---
title: "Stats for Sociologists"
subtitle: "SOC6302H1F"
format:
  html:
    toc: true
    output-file: index
  pdf:
    toc: false
    number-sections: true
    number-depth: 3
    geometry:
      - top=2cm
      - bottom=3cm
      - left=2.5cm
      - right=2.5cm
  docx: 
    toc: false
    number-sections: true
    number-depth: 3
    geometry:
      - top=2cm
      - bottom=3cm
      - left=2.5cm
      - right=2.5cm
block-headings: false
filters:
    - openlinksinnewpage
---

```{r, include=FALSE}

## Install packages not yet installed & load them
pacman::p_load(
  tidyverse,
  knitr,
  flextable, # formatted pretty tables
  googledrive,
  googlesheets4,
  tinytex,
  conflicted # declare primary packages
)

## sheets reauth with specified token and email address
drive_deauth()
googledrive::drive_auth(
  cache = ".secrets",
  email = "joannapepin@gmail.com"
)

# Give google sheets the same authorization
gs4_auth(token = drive_token())
```

Fall 2025

**Schedule:** Tuesdays, 10AM - 12PM (Lab 12PM - 1PM)

**Instructor:** Joanna R. Pepin \[j.pepin\@utoronto.ca\]

**Teaching Assistant:** Firdaous Sbaï
\[firdaous.sbai\@mail.utoronto.ca\]

**Office Hours:** By appointment

# Course Description

This course introduces students to descriptive and basic inferential
statistical techniques, focusing on practical applications of analyzing
quantitative data through computing. Laying the groundwork for more
advanced classes students may take in the future, the course covers key
concepts such as summary statistics, sampling distributions, hypothesis
testing, correlation, and linear regression – all framed through the
lens of real-world applications. Learning is hands-on, using R
statistical software to compute and interpret these techniques with
publicly available data sets commonly used in social science research.
Though self-paced online tutorials, interactive lectures, and lab
sessions, students will gain the skills needed to analyze social science
data and draw evidence-based conclusions. Students will also practice
communicating statistical findings.

## Course Objectives

More specifically, students who successfully complete this course should
be able to…

-   demonstrate introductory knowledge of statistical programming.
-   select and compute descriptive and basic inferential statistics
    using R.
-   read and interpret basic statistics as they appear in academic
    publications.
-   communicate statistical methods and findings through reproducible
    reports and oral presentations.

# Course Materials and Resources

**MATERIALS**

There is no required textbook for this course.

You will need to **bring a laptop computer to class**, as you are
expected to interact and program throughout the lecture sessions and the
lab.

**SOFTWARE**

The statistical programming language for this course will be
[R](https://www.r-project.org/about.html). Computations are executed
from a set of typed commands which is easiest to undertake through
[RStudio](https://rstudio.com/products/rstudio/), an editor which allows
you to see your code and output. Both R and RStudio are free to download
to your own personal computer:

1.  Download [R](https://cran.rstudio.com/)
2.  Download [RStudio (free
    version)](https://rstudio.com/products/rstudio/download/)

You will also need regular access to strong, reliable internet for this
class.

# Course Design

This course consists of three components: a self-paced online tutorial
delivered asynchoronously, a two-hour in-person interactive lecture, and
a one-hour lab.

**Online Tutorial:** Before most classes, there will be an online
tutorial to complete at your own pace. The link to the tutorial will be
available on Quercus. The asyhcnonous online tutorials will contain
self-check questions (not for credit).

It is easy to get lost in a course where each week’s content builds on
the material of previous weeks. It is crucial to keep up with the
tutorials regularly to help you stay on track.

**Interactive Lecture**: The in-person interactive lecture consists of
two parts. In the first part, I will ask and solicit questions about the
week’s online tutorial, focusing on any points of confusion or tricky
components. In the second part, I will demonstrate data analysis using R
and RStudio. Make sure to bring a laptop computer, as you are expected
to interact and program along.

**Lab:** The one-hour lab session is self-led, with TA guidance and
assistance available. This is a supportive environment in which you are
expected to work on the upcoming milestone and your individual research
briefs. The TA will actively provide help on coding and interpretation,
giving key tips and advice during the lab. If you have already completed
the upcoming milestone, then you are expected to help other students
during the lab or work on your research brief (attendance is mandatory).

## Components

All assignments, evaluations, and tests must be submitted via Quercus.

```{r echo = FALSE}
Assessment <- c("Milestones (MS) 1-4", "Midterm", 
                "Flash Presentation", "Research Brief")
Worth <- c("40%", "20%", "15%", "25%")
Due <- c("Variable", "October 21", "TBD", "November 25")

summary <- data.frame(Assessment, Worth, Due)
```

```{r echo = FALSE}
flextable(summary) %>%
  set_table_properties(width = 1, layout = "autofit") %>%
  theme_zebra() %>%
  bg(bg = "#18BC9C", part = "header") %>%
  color(color = "white", part = "header")
```

# Schedule

*Topics and topic order are subject to changes.*

```{r message =FALSE, echo=FALSE}
sched <- read_sheet("1W3SSN_HG9hULeiVC66upl57uxtXntY96xCs9ru5yWe4", 
                    sheet = "Schedule")
```

```{r, echo = FALSE}
flextable(sched) %>%
  set_table_properties(width = 1, layout = "autofit") %>%
  theme_zebra() %>%
  bg(bg = "#18BC9C", part = "header") %>%
  color(color = "white", part = "header") %>%
  color(color = "#E74C3C", i = 8, j = "Tutorial") %>%
  bold(bold = TRUE, i = 8, j = "Tutorial") %>%
  color(color = "#18BC9C", i = 9, j = "Tutorial") %>%
  bold(bold = TRUE, i = 9, j = "Tutorial") %>%
  color(color = "#E74C3C", i = 13, j = "Tutorial") %>%
  bold(bold = TRUE, i = 13, j = "Tutorial") %>%
  italic(italic = TRUE, i = 14, j = "Dates")
```

# Evaluation

## Milestones (40%)

Four milestone (MS) assignments will be assigned throughout the
semester. Each MS will be weighted equally (10% each).

Milestones (MS) will include a mixture of conceptual and applied
(computational) questions. Sometimes a MS will incorporate tasks that
require you to demonstrate progress on your flash presentations or
research brief.

Milestones should be completed using Quarto, and the submission should
include the knitted pdf and Quarto file containing the necessary code to
produce the results in your MS.

Milestones are due by the beginning of class on the specified date. They
should be uploaded to the designated spot on Quercus (NOT emailed to
your professor or TA).

You may work together in small groups on milestone assignments.
Remember, however, that working through the milestone is the best
preparation for the exam and a measure of your understanding of course
material. Even if you work with others, you must write up and turn in
your own MS.

## Midterm (20%)

The midterm will consist of multiple-choice questions assessing
conceptual understanding of the course material and may involve some
basic calculations by hand. You will not use R during the exams;
however, you will be asked to interpret R coding syntax and output.

## Flash Presentation (15%)

You will give a short (approximately 5 minutes) presentation on your
research brief. You will be required to submit the slides (about 5
slides) for your presentation to Quercus. Details on the flash
presentation will be given at a later date.

## Research Brief (25%)

You will apply the skills you’ve been learning to answer a research
question that is of interest to you. Your Research Brief should include:

-   **Question**: a clearly worded research question that can be
    answered with quantitative data.
-   **Data**: a description of the data you are using, including a
    description of all variables and how they are coded. A table of
    descriptive statistics of the total sample and key comparison groups
    and a narrative of the table.
-   **Plan of Analysis**: a description of your analysis (i.e., how are
    you going to use the data to answer the research question?)
-   **Results**: a presentation of the results, including a regression
    table and a figure.
-   **Discussion**: Provide an interpretation of the results – that is,
    given the results, what is the answer to the question? Are there any
    other possible interpretations? What limitations might there be in
    your data or analysis that affect how well we can answer the
    question?

Details on the research brief, including choice of datasets, will be
given at a later date.

**NOTE:** To keep you on track and to provide feedback along the way,
you will submit Table 01 (including your research question and data
section) earlier in the semester (5%). Your complete research brief
submitted at the end of the semester will therefore be 20% of your final
grade.

## Grading Scale

Final grades will adhere to the [Arts and Sciences Graduate Grade
Scale](https://www.registrar.utoronto.ca/records-academics/transcripts/grading-scales-notations/).

| Grade | Score | Meaning |
|------------------------|------------------------|------------------------|
| Excellent | 5 | Strong evidence of original thinking; good organization; capacity to analyze and synthesize; superior grasp of course material with sound critical evaluations. |
| Good | 4 | Evidence of grasp of course material, some evidence of critical capacity and analytic ability; reasonable evidence of critical and analytic skills. |
| Adequate | 3 | Evidence of familiarity with course material and some evidence that critical and analytic skills have been developed. |
| Marginal | 2 | Some evidence of familiarity with the subject matter and weakness in critical and analytic skills; |
| Inadequate | 1 | Little evidence of even superficial understanding of subject matter; limited critical and analytic skills; irrelevant use of course material. |

: Research briefs will be graded as follows:
{tbl-colwidths="\[25,10,65\]"}

# Course Policies

## Assignment Submission

All written work must be typed, double-spaced, with 1-inch margins using
11- or 12-point Times New Roman, Aptos, Calibri, or Roboto.

Written work must be submitted via Quercus. No work will be accepted
over e-mail. Files should be in PDF or .doc(x) format or as a .R script
if requested. Other file formats will not be graded and will be treated
as unsubmitted assignments. It is your responsibility to ensure that the
file itself is readable, which can be confirmed by downloading the file
once it is uploaded to the assignment submission page on Quercus.
Unreadable files are missed assignments.

## Late Policy & Course Extensions

Falling behind in a statistics course generally makes learning
exponentially more difficult and thus, late work is generally not
accepted.

However, a due date for a Milestone assignment or the Midterm may be
extended for a legitimate reason with proper documentation (e.g.,
illness, family emergency, religious observance, accessibility
accommodation) according to UofT's guidelines.

Late Research Briefs will be penalized at a rate of 2% per day up to 10%
(5 days) and will not be accepted thereafter, except where there are
legitimate, documented reasons beyond a student’s control.

Flash Presentations, given their in class participatory nature, cannot
be extended.

***Note:*** Extensions beyond the term end date (e.g., the last day of a
fall term class may be December 1st, but the term ends December 23rd)
requires a discussion with the instructor and the graduate office, as
well as completion of an SGS request for an extension of course work
form. These forms will be considered by the graduate office and are not
automatically approved.

## Academic Integrity

Students are expected to know and adhere to the University’s principles
of academic integrity. Copying, plagiarizing, falsifying medical
certificates, or other forms of academic misconduct will not be
tolerated. Any act of plagiarism or other unethical behavior will be
addressed in accordance with University guidelines. Please see the [Code
of Behaviour on Academic
Matters](http://www.governingcouncil.utoronto.ca/policies/behaveac.htm)
for specific information on academic integrity at UofT.

## Plagiarism Detection

Turnitin plagiarism detection software is integrated within Quercus. It
uses text matching technology as a method to uphold the University’s
high academic integrity standards to detect any potential plagiarism.
The terms that apply to the University’s use of this tool are described
on the [Centre for Teaching Support &
Innovation](https://uoft.me/pdt-faq.) web site. I am not comfortable
with the widespread use of this software, which effectively provides
large quantities of student work to a for-profit company whose future
business uses may include using it to train machine learning and AI
tools. Consequently, none of the assignments for this course are set for
automatic review by Turnitin. However, if I have reason to suspect
plagiarism, I may use this tool to detect plagiarism.

If you object to your work being subject to review by Turnitin, add a
comment when you submit your assignment. If I do find that we need to
review your assignment for plagiarism and you have asked that I not use
Turnitin, I will ask that you provide sufficient secondary material
(e.g., reading notes, outlines of the paper, rough drafts of the final
draft, etc.) to establish that the paper you submit is truly yours. If
you cannot or decline to provide the requested documentation and do not
consent to us using Turnitin for review, I will decline to mark the
assignment.

**Plagiarism:** Be careful to avoid plagiarism. It is a serious academic
offense with serious penalties (see the [Code of Behavior on Academic
Matters](https://governingcouncil.utoronto.ca/secretariat/policies/code-behaviour-academic-matters-july-1-2019)).
Do not present someone else’s ideas as your own. Give proper references
to others’ ideas, and use quotation marks if you are quoting. When in
doubt, err on the side of a reference. Turning in an old paper, or large
parts thereof, for credit in a second (or third etc.) course, is
considered an academic offense that results in students being referred
off to the Office of Academic Integrity.

## Artificial Intelligence

Students may choose to use generative artificial intelligence (AI) tools
as they work through the assignments. I encourage you to limit your use
to technical coding support/reference, grammar/copy editing, and
outlining. Be aware that AI tools are often wrong, overly generic, and
lacking in critical nuance. My expectations for you are considerably
higher than that. You will need to be more persuasive and creative at
linking sociological research with empirical data, thinking critically,
and making connections between data and theory than any of these tools.
Over-reliance on AI will prevent you from learning the skills necessary
to complete work at the level expected for this class. Students are
ultimately accountable for the work they submit.

## Accessibility and Student Accommodations

Students with diverse learning styles and needs are welcome in this
course. The University of Toronto is committed to accessibility. If you
require accommodations or have any accessibility concerns, please visit
<http://studentlife.utoronto.ca/as> as soon as possible.

## Equity and Diversity Statement

The University of Toronto is committed to equity and respect for
diversity. All members of the learning environment in this course should
strive to create an atmosphere of mutual respect. As a course
instructor, I will neither condone nor tolerate behaviour that
undermines the dignity or self-esteem of any individual in this course
and wish to be alerted to any attempt to create an intimidating or
hostile environment. It is our collective responsibility to create a
space that is inclusive and welcomes discussion. Discrimination,
harassment and hate speech will not be tolerated. Additional information
and reports on Equity and Diversity at the University of Toronto is
available at
[http://equity.hrandequity.utoronto.ca](http://equity.hrandequity.utoronto.ca/).
