---
title: "SOC6302 data"
format: html
editor: visual
---

## Install packages not yet installed & load them

```{r}
pacman::p_load(
  tidyverse,
  gssr, # load U.S. gss data
  gssrdoc, # load the gss codebook
  haven, # work with labeled data
  labelled, # use labels as values
  summarytools, # easy descriptive statistics
  conflicted # declare primary packages
)
```

## Address important conflicts

```{r}
conflict_prefer("is.labelled", "haven")
conflict_prefer("filter", "dplyr")
conflict_scout() # Identify the conflicts
```

## Load the gss datasets

```{r}
# Load all gss
data(gss_all)

# Get the data only for the 2024 survey respondents
gss24 <- gss_get_yr(2024)

# Load the codebook
data(gss_dict)
```

# 01.1 Intro

# Lecture 01

# 02.1 Terms

**frequency:** `premarsx`

```{r}
# GSS_ALL
gss_all |>
  select(premarsx, sex) |>
  saveRDS(file = "tutorials/SOC6302-02/data/gss_all.rds")

# GSS24
gss24 |>
  select(premarsx, sex, fefam) |>
  saveRDS(file = "tutorials/SOC6302-02/data/gss24.rds")
```

# Lecture 02

**haven labelled:** `premarsx`; `sex`\
**frequency:** `premarsx`;\
**crosstab:** `premarsx`;

# Milestone 01

# 03.1 Descriptives

**mean:** fake data\
**median:** `wkvsfam`; `agekdbrn` (histogram);\
**mode:** `helpful;` `fairern`;\
**distribution shape:** `agekdbrn`;

```{r}
## GSS_ALL
gss_all |>
  select(year, wkvsfam, agekdbrn, fairearn, socrel) |>
  saveRDS(file = "tutorials/SOC6302-03/data/gss_all.rds")

## GSS24
gss24 |>
  select(helpful, premarsx) |>
  saveRDS(file = "tutorials/SOC6302-03/data/gss24.rds")
```

# Lecture 03

**dplyr():** `agekdbrn`; `sex;`

maybe use agekdbrn for mean median and mode? practice manipulating into
categories, then back to numeric?

**mode:**

**median:**

**mean:** `hrs1`;

**Think Like a Statistician:** `premarsx`; `polviews`

# Milestone 02

# 04.1 Sampling

```{r}
# GSS_ALL
gss_all |>
  select(year, agekdbrn) |>
  drop_na() |>
  subset(year == 2022) |>
  saveRDS(file = "tutorials/SOC6302-04/data/gss_all.rds")
```

# Lecture 04
