---
title: "Stats for Sociologists"
subtitle: "SOC6302H1F"
format:
  html:
    toc: true
    output-file: index
  pdf:
    toc: false
    number-sections: true
    number-depth: 3
    geometry:
      - top=2cm
      - bottom=3cm
      - left=2.5cm
      - right=2.5cm
  docx: 
    toc: false
    number-sections: true
    number-depth: 3
    geometry:
      - top=2cm
      - bottom=3cm
      - left=2.5cm
      - right=2.5cm
block-headings: false
filters:
    - openlinksinnewpage
---

```{r, include=FALSE}

## Install packages not yet installed & load them
pacman::p_load(
  tidyverse,
  knitr,
  flextable, # formatted pretty tables
  googledrive,
  googlesheets4,
  tinytex,
  conflicted # declare primary packages
)

## sheets reauth with specified token and email address
drive_deauth()
googledrive::drive_auth(
  cache = ".secrets",
  email = "joannapepin@gmail.com"
)

# Give google sheets the same authorization
gs4_auth(token = drive_token())
```

Fall 2026

**Schedule:** Thursdays, 10AM - 12PM (Lab: Fridays, 11AM - 12PM)

**Instructor:** Joanna R. Pepin \[j.pepin\@utoronto.ca\]

**Teaching Assistant:** Someone Great
\[someone.great\@mail.utoronto.ca\]

**Office Hours:** By appointment

# Course Description

This course introduces students to descriptive and basic inferential
statistical techniques, focusing on practical applications of analyzing
quantitative data through computing. Laying the groundwork for more
advanced classes students may take in the future, the course covers key
concepts such as summary statistics, sampling distributions, hypothesis
testing, correlation, and linear regression – all framed through the
lens of real-world applications. Learning is hands-on, using R
statistical software to compute and interpret these techniques with
publicly available data sets commonly used in social science research.
Though self-paced online tutorials, interactive lectures, and lab
sessions, students will gain the skills needed to analyze social science
data and draw evidence-based conclusions. Students will also practice
communicating statistical findings.

## Course Objectives

More specifically, students who successfully complete this course should
be able to…

- demonstrate introductory knowledge of statistical programming.
- select and compute descriptive and basic inferential statistics
  using R.
- read and interpret basic statistics as they appear in academic
  publications.
- communicate statistical methods and findings through reproducible
  reports and oral presentations.

# Course Materials and Resources

**MATERIALS**

There is no required textbook for this course.

You will need to **bring a laptop computer to class**, as you are
expected to interact and program throughout the lecture sessions and the
lab.

**SOFTWARE**

The statistical programming language for this course will be
[R](https://www.r-project.org/about.html). Computations are executed
from a set of typed commands which is easiest to undertake through
[RStudio](https://rstudio.com/products/rstudio/), an editor which allows
you to see your code and output. Both R and RStudio are free to download
to your own personal computer:

1.  Download [R](https://cran.rstudio.com/)
2.  Download [RStudio (free
    version)](https://rstudio.com/products/rstudio/download/)

You will also need regular access to strong, reliable internet for this
class.

# Course Design

This course consists of three components: a self-paced online tutorial
delivered asynchoronously, a two-hour in-person interactive lecture, and
a one-hour lab.

**Online Tutorial:** Before most classes, there will be an online
tutorial to complete at your own pace. The link to the tutorial will be
available on Quercus. The asyhcnonous online tutorials will contain
self-check questions (not for credit).

It is easy to get lost in a course where each week’s content builds on
the material of previous weeks. It is crucial to keep up with the
tutorials regularly to help you stay on track.

**Interactive Lecture**: During the in-person interactive lecture, I
will demonstrate data analysis using R and RStudio, applying the
concepts of the online tutorial. Make sure to bring a laptop computer,
as you are expected to interact and program along. Throughout the class,
I will ask and solicit questions about the week’s online tutorial,
focusing on any points of confusion or tricky components.

**Lab:** The one-hour lab session will begin with a brief TA-led group
activity focused on practice exercises drawn from the material covered
in the online tutorial and lecture. The majority of the session will be
self-directed, during which you are expected to work on the upcoming
milestone and your individual research brief. This is a supportive,
collaborative environment in which students are expected to work
independently while also learning from and supporting one another. The
TA will circulate throughout the lab to provide guidance, answer
questions, and assist with coding and interpretation as needed. If you
finish the milestone early, you are encouraged to help your peers or
continue working on your research brief (attendance is mandatory).

## Components

All assignments, evaluations, and tests must be submitted via Quercus.

```{r echo = FALSE}
Assessment <- c("Milestones (MS) 1-4", "Midterm", 
                "Flash Presentation", "Research Brief")
Worth <- c("30%", "25%", "15%", "30%")
Due <- c("Variable", "October 21", "November 25","TBD")

summary <- data.frame(Assessment, Worth, Due)
```

```{r echo = FALSE}
flextable(summary) |>
  set_table_properties(width = 1, layout = "autofit") |>
  theme_zebra() |>
  bg(bg = "#18BC9C", part = "header") |>
  color(color = "white", part = "header")
```

# Schedule

*Topics and topic order are subject to changes.*

```{r message =FALSE, echo=FALSE}
sched <- read_sheet("1W3SSN_HG9hULeiVC66upl57uxtXntY96xCs9ru5yWe4", 
                    sheet = "Schedule")
```

```{r, echo = FALSE}
flextable(sched) |>
  set_table_properties(width = 1, layout = "autofit") |>
  theme_zebra() |>
  bg(bg = "#18BC9C", part = "header") |>
  color(color = "white", part = "header") |>
  color(color = "#E74C3C", i = 6, j = "Lecture") |>
  bold(bold = TRUE, i = 6, j = "Lecture") |>
  color(color = "#E74C3C", i = 12, j = "Lecture") |>
  bold(bold = TRUE, i = 12, j = "Lecture") |>
  italic(italic = TRUE, i = 13, j = "Dates")
```

# Evaluation

## Milestones (30%)

Four milestone (MS) assignments will be assigned throughout the
semester. MS01 and MS04 are each worth 5%, MS02 and MS03 are each worth
10%.

Milestones will include a mixture of conceptual and applied
(computational) questions. Sometimes a MS will incorporate tasks that
require you to demonstrate progress on your flash presentations or
research brief.

Milestones will be completed using Quarto. The submission must include
the knitted pdf AND Quarto file containing the necessary code to produce
the results in your MS.

Milestones are due by the beginning of class on the specified date. They
should be uploaded to the designated spot on Quercus (NOT emailed to
your professor or TA).

You may work together in small groups on milestone assignments. Sharing
code, discussing interpretations, and helping one another understand
concepts are encouraged. However, all submitted work must reflect your
own understanding and effort. Writing code or answers for another
student, asking someone else to write them for you, or submitting work
that you do not understand or could not explain yourself is a violation
of academic integrity. Completing the milestones yourself is not only
the best preparation for this course (i.e. the midterm and research
brief), but also for developing the skills needed to work as an
independent scholar beyond the classroom.

## Midterm (25%)

The midterm will consist of multiple-choice questions assessing
conceptual understanding of the course material and may involve some
basic calculations by hand. You will not use R during the exams;
however, you will be asked to interpret R coding syntax and output.

## Flash Presentation (15%)

You will give a short (approximately 5 minutes) presentation on your
research brief. You will be required to submit the slides (about 5
slides) for your presentation to Quercus. Details on the flash
presentation will be given at a later date.

## Research Brief (30%)

You will apply the skills you’ve been learning to answer a research
question that is of interest to you. Your Research Brief should include:

- **Question**: a clearly worded research question that can be answered
  with quantitative data.
- **Data**: a description of the data you are using, including a
  description of all variables and how they are coded. A table of
  descriptive statistics of the total sample and key comparison groups
  and a narrative of the table.
- **Plan of Analysis**: a description of your analysis (i.e., how are
  you going to use the data to answer the research question?)
- **Results**: a presentation of the results, including a regression
  table and a figure.
- **Discussion**: Provide an interpretation of the results – that is,
  given the results, what is the answer to the question? Are there any
  other possible interpretations? What limitations might there be in
  your data or analysis that affect how well you can answer the
  question?

Details on the research brief, including choice of datasets, will be
given at a later date.

**NOTE:** To help keep you on track for your final project and receive
feedback along the way, each Milestone will include a small portion
related to your project–some will serve as stepping stones to guide your
thinking, while others will be components that may directly contribute
to your final submission.

## Grading Scale

Final grades will adhere to the [Arts and Sciences Graduate Grade
Scale](https://www.registrar.utoronto.ca/records-academics/transcripts/grading-scales-notations/).

Research briefs will be graded as follows:

| Grade | Meaning |
|-------------|------------------------------------------|
| Excellent | Strong evidence of original thinking; good organization; capacity to analyze and synthesize; superior grasp of course material with sound critical evaluations. |
| Good | Evidence of grasp of course material, some evidence of critical capacity and analytic ability; reasonable evidence of critical and analytic skills. |
| Adequate | Evidence of familiarity with course material and some evidence that critical and analytic skills have been developed. |
| Marginal | Some evidence of familiarity with the subject matter and weakness in critical and analytic skills. |
| Inadequate | Little evidence of even superficial understanding of subject matter; limited critical and analytic skills; irrelevant use of course material. |

# Course Policies

## Assignment Submission

All written work must be typed, double-spaced, with 1-inch margins using
11- or 12-point Times New Roman, Aptos, Calibri, or Roboto.

Written work must be submitted via Quercus. No work will be accepted
over e-mail. Files should be in PDF or .doc(x) format or as a .R script
if requested. Other file formats will not be graded and will be treated
as unsubmitted assignments. It is your responsibility to ensure that the
file itself is readable, which can be confirmed by downloading the file
once it is uploaded to the assignment submission page on Quercus.
Unreadable files are missed assignments.

## Late Policy & Course Extensions

Falling behind in a statistics course generally makes learning
exponentially more difficult and thus, late work is generally not
accepted.

However, a due date for a Milestone assignment or the Midterm may be
extended for a legitimate reason with proper documentation (e.g.,
illness, family emergency, religious observance, accessibility
accommodation) according to UofT's guidelines.

Late Research Briefs will be penalized at a rate of 2% per day up to 10%
(5 days) and will not be accepted thereafter, except where there are
legitimate, documented reasons beyond a student’s control.

Flash Presentations, given their in class participatory nature, cannot
be extended.

***Note:*** Extensions beyond the term end date (e.g., the last day of a
fall term class may be December 1st, but the term ends December 23rd)
requires a discussion with the instructor and the graduate office, as
well as completion of an SGS request for an extension of course work
form. These forms will be considered by the graduate office and are not
automatically approved.

## Academic Integrity

Students are expected to know and adhere to the University’s principles
of academic integrity. Copying, plagiarizing, falsifying medical
certificates, or other forms of academic misconduct will not be
tolerated. Any act of plagiarism or other unethical behavior will be
addressed in accordance with University guidelines. Please see the [Code
of Behaviour on Academic
Matters](http://www.governingcouncil.utoronto.ca/policies/behaveac.htm)
for specific information on academic integrity at UofT.

## Plagiarism Detection

Turnitin plagiarism detection software is integrated within Quercus. It
uses text matching technology as a method to uphold the University’s
high academic integrity standards to detect any potential plagiarism.
The terms that apply to the University’s use of this tool are described
on the [Centre for Teaching Support &
Innovation](https://uoft.me/pdt-faq.) web site. I am not comfortable
with the widespread use of this software, which effectively provides
large quantities of student work to a for-profit company whose future
business uses may include using it to train machine learning and AI
tools. Consequently, none of the assignments for this course are set for
automatic review by Turnitin. However, if I have reason to suspect
plagiarism, I may use this tool to detect plagiarism.

If you object to your work being subject to review by Turnitin, add a
comment when you submit your assignment. If I do find that we need to
review your assignment for plagiarism and you have asked that I not use
Turnitin, I will ask that you provide sufficient secondary material
(e.g., reading notes, outlines of the paper, rough drafts of the final
draft, etc.) to establish that the paper you submit is truly yours. If
you cannot or decline to provide the requested documentation and do not
consent to us using Turnitin for review, I will decline to mark the
assignment.

**Plagiarism:** Be careful to avoid plagiarism. It is a serious academic
offense with serious penalties (see the [Code of Behavior on Academic
Matters](https://governingcouncil.utoronto.ca/secretariat/policies/code-behaviour-academic-matters-july-1-2019)).
Do not present someone else’s ideas as your own. Give proper references
to others’ ideas, and use quotation marks if you are quoting. When in
doubt, err on the side of a reference. Turning in an old paper, or large
parts thereof, for credit in a second (or third etc.) course, is
considered an academic offense that results in students being referred
off to the Office of Academic Integrity.

## Artificial Intelligence

Students are strongly discouraged from using generative AI for coding,
statistical interpretation, or substantive assignment work. AI can
sometimes be useful for simple technical questions, but it is often
incorrect in ways that can be difficult to evaluate if you are still
learning the material. A key objective of this course is to develop your
ability to troubleshoot problems, interpret results, and find reliable
information. Before turning to AI, make use of the online tutorials,
lecture slides, and your peers and TA, all of which will often lead to a
faster and deeper understanding of the material and stronger long-term
skills. Students are ultimately responsible for all work they submit and
are particularly encouraged to complete their own interpretation of
results.

Students suspected of over reliance on AI may be required to meet with
the instructor for an oral assessment of a milestone or research brief
to demonstrate their understanding of the submitted work.

## Accessibility and Student Accommodations

The University of Toronto is committed to accessibility. If you require
accommodations or have any accessibility concerns, please visit
<http://studentlife.utoronto.ca/as> as soon as possible.

## Equity and Diversity Statement

The University of Toronto is committed to equity and respect for
diversity. All members of the learning environment in this course should
strive to create an atmosphere of mutual respect. As a course
instructor, I will neither condone nor tolerate behaviour that
undermines the dignity or self-esteem of any individual in this course
and wish to be alerted to any attempt to create an intimidating or
hostile environment. It is our collective responsibility to create a
space that is inclusive and welcomes discussion. Discrimination,
harassment and hate speech will not be tolerated. Additional information
and reports on Equity and Diversity at the University of Toronto is
available at
[http://equity.hrandequity.utoronto.ca](http://equity.hrandequity.utoronto.ca/).
