---
title: "Statistical Tests I"
format:
  revealjs:
    multiplex: false
    controls: true
    theme: [default, custom.scss]
    highlight-style: atom-one
    include-in-header:
    - file: mathjax-color.html
    slide-number: c
    math: mathjax
    code-annotations: hover
revealjs-plugins:
  - quiz
embed-resources: true
filters:
    - openlinksinnewpage
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: true
---

```{r echo=FALSE}

#| label: setup
#| message: false
#| include: false

## Load packages, custom functions, and styles
source("custom-setup-styles.R")

# Color in math equations
# https://egallic.fr/Recherche/aide-memoire/quarto_equation_colors.html#colors-in-equations-with-revealjs
```

# Agenda {data-menu-title="Agenda" background-image="images/calendar.png" background-size="100px" background-repeat="repeat"}

-   X
-   Y

::: notes
-  missing data? 
-  **outliers** 
-  dates? 
-  lists
-  forcats (reorder factor levels)
:::

# Learning objectives

By the end of the lecture, you will be able to ...

-   X 
-   Y
-   Z

# Code-along 05 {data-menu-title="Code-along" background-image="images/terminal.png"}

Download and open [code-along-05.qmd]()

## Packages {.sparse-slide}

Load the standard packages.

```{r}
#| message: false
#| label: load

library(here)
library(tidyverse) 
library(haven) # not core tidyverse
library(gssr)
library(gssrdoc)
library(summarytools)
```

## Load your data & codebook

```{r}
#| message: false
#| label: gssdata

# Get the data all survey years
data(gss_all)

# Load the codebook
data(gss_dict)

```

# X {.theme-section}
# Y {.theme-section}
