---
title: "Summarizing Data"
format:
  revealjs:
    multiplex: false
    controls: true
    theme: [default, custom.scss]
    highlight-style: atom-one
    include-in-header:
    - file: mathjax-color.html
    slide-number: c
    math: mathjax
    code-annotations: hover
revealjs-plugins:
  - quiz
embed-resources: true
filters:
    - openlinksinnewpage
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: true
---

```{r echo=FALSE}

#| label: setup
#| message: false
#| include: false

## Load packages, custom functions, and styles
source("custom-setup-styles.R")

# Color in math equations
# https://egallic.fr/Recherche/aide-memoire/quarto_equation_colors.html#colors-in-equations-with-revealjs
```

# Agenda {data-menu-title="Agenda" background-image="images/calendar.png" background-size="100px" background-repeat="repeat"}

-   Data Management II
-   Central Tendency & Measures of Variability

# Learning objectives

By the end of the lab, you will be able to ...

-   produce measures of central tendency and variability

# Code-along 03 {.theme-section}

Download and open [code-along-03.qmd]()




# Central Tendency & Variability {.theme-section}

## Mode

Remember, the **mode** is the category with the greatest frequency (or
the largest percentage). Let's find it for the `premarsx` variable.

<br>

```{r}
#| label: mode

freq(gss24$premarsx, report.nas = FALSE) 

```

## Median

We can use the same table we generated before to identify the
**median**. This time, let's use [dplyr grammar]() to produce the same
table

Remember, use the cumulative percentage to locate the **50th
percentile**.

```{r}
#| label: median
#| code-line-numbers: "1|2|3"

gss24 |> # <1>
  freq(premarsx, report.nas = FALSE) |> # <2>
  tb() # <3> 

```

1.  Use `dplyr` grammar, starting with the name of the df and a pipe
2.  Use the `freq()` function as usual
3.  Add the `tb()` function to turn the table into a `tibble`

## Mean

```{r}
#| label: freq-numeric

freq(gss24$hrs1, report.nas = FALSE, headings = FALSE) 

```

::: aside
Question 1a. If working, full or part time: how many hours did you work
last week, at all jobs?
:::

## Mean

```{r}
#| label: mean-median

mean(gss24$hrs1, na.rm=TRUE) # <1>
median(gss24$hrs1, na.rm=TRUE)

```

1.  `na.rm` is a logical evaluating to TRUE or FALSE indicating whether
    NA values should be stripped before the computation proceeds.

::: aside
Question 1a. If working, full or part time: how many hours did you work
last week, at all jobs?
:::

## `summary()`

<br>

```{r}
#| label: summary

summary(gss24$hrs1)

```

## Variability: `descr()`

<br>

```{r}
#| label: desc01

descr(gss24$hrs1)

```

## Variability: `descr()`

```{r}
#| label: descr02
#| code-line-numbers: "1|2|3|4"

gss24 |> 
  descr(hrs1,
        stats = "common") |> # <1>
  tb() 

```

1.  Which stats to produce. Either “all” (default), “fivenum”, “common”
    (see Details), or a selection. See ?descr