---
title: "Summarizing Data"
format:
  revealjs:
    multiplex: false
    controls: true
    theme: [default, custom.scss]
    highlight-style: atom-one
    include-in-header:
    - file: mathjax-color.html
    slide-number: c
    math: mathjax
    code-annotations: hover
revealjs-plugins:
  - quiz
embed-resources: true
filters:
    - openlinksinnewpage
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: true
---

```{r echo=FALSE}

#| label: setup
#| message: false
#| include: false

## Load packages, custom functions, and styles
source("custom-setup-styles.R")

# Color in math equations
# https://egallic.fr/Recherche/aide-memoire/quarto_equation_colors.html#colors-in-equations-with-revealjs
```

# Agenda {data-menu-title="Agenda" background-image="images/calendar.png" background-size="100px" background-repeat="repeat"}

-   Data Management I
-   Central Tendency & Measures of Variability

# Learning objectives

By the end of the lab, you will be able to ...

-   implement basic variable manipulation
-   produce measures of central tendency and variability

# Code-along 03 {.theme-section}

Download and open [code-along-03.qmd]()

## Packages

Load the standard packages.

```{r}
#| message: false
#| label: load

library(here)
library(tidyverse) 
library(haven) # not core tidyverse
library(gssr)
library(gssrdoc)
library(summarytools)
```

## Load your data & codebook

```{r}
#| message: false
#| label: gssdata

# Get the data all survey years
data(gss_all)

# Load the codebook
data(gss_dict)

```


# Data Management I {.theme-section}

## Tidying Data {.smaller}

::: {style="color: #18bc9c; font-size: 150%; font-family: 'Shadows Into Light'"}
Most tasks related to data analysis are not glorious or fancy.  
:::

<br>
  
A lot of time is dedicated to whipping a dataset into the shape needed to be able to analyze it.  
 
<br>
  
This task has different names “data cleaning,” “data management,” “data manipulation,” “data wrangling,” “data transformation.”  

::: aside
Source: [Jasper Dag Tjaden](https://jaspertjaden.github.io/course-intro2r/week3.html)
:::

## `dplyr` package {.smaller}

The `dplyr` package provides a complete set of functions that help you solve the most common data manipulation challenges such as:  

-    filtering observations based on their values
-    extracting observations based on their values or positions
-    sampling observations based on a specific number or fraction of rows
-    sorting observations based on one or several variables
-    selecting variables based on their names or positions
-    renaming variables
-    adding new variables based on existing ones
-    summarizing observations or variables to a single descriptive measure
-    performing any operation by group

::: aside
Source: [Antoine Soetewey](https://statsandr.com/blog/introduction-to-data-manipulation-in-r-with-dplyr/)
:::

## The pipe `|>` 

<br>

The pipe operator passes what comes before it into the function that comes after it as the first argument in that function.

```{r}
sum(1, 2)
```

<br>

. . . 

```{r}
1 |> 
  sum(2)
```


## `dplyr` style {.smaller}

In data transformation pipelines, always use a  

-   space before `|>` 
-   line break after `|>` 
-   indent the next line of code 
    
::: {.smaller}
We'll talk about data visualization pipes later...
:::

<br>
    
::: {.callout-tip icon="false"}
## {{< fa hat-wizard >}} Heads Up!

|> (native pipe operator) and %>% (magrittr package) behave identically for simple cases. [More info.]((https://www.tidyverse.org/blog/2023/04/base-vs-magrittr-pipe/))
:::

## `function`(argument) 

Functions are (most often) verbs, followed by what they will be applied
to in parentheses:

<br>

```{r}
#| eval: false
#| message: false
#| label: function

do_this(to_this)
do_that(to_this, to_that, with_those)
```

<br>

`dplyr` verbs (functions) will allow you to solve the vast majority of your data manipulation challenges.  

::: aside
Source: [Dr. Mine Çetinkaya-Rundel](https://sta199-f24.github.io/slides/01-meet-the-toolkit-slides.html#/functionargument)  
:::


## `dplyr` basics 

::: {.smaller}
They are organized into four groups based on what they operate on: rows, columns, groups, or tables.  
:::

<br>
  
**The verbs all have in common**:  

1.   The first argument is always a data frame. 
2.   The subsequent arguments typically describe which columns to operate on using the variable names (without quotes). 
3.   The output is always a new data frame. 
 
::: aside
Source: [R for Data Science](https://r4ds.hadley.nz/data-transform.html#the-pipe)
:::

## `dplyr` grammar

What's the advantage of
[`dplyr`](https://cran.r-project.org/web/packages/dplyr/readme/README.html)
grammar? We can sequence data manipulation!

<br>


```{r}
#| label: pipe00
#| output-location: column

gss_all |> 
  select(year, sex, agekdbrn) |>
  filter(year == 2022) |>
  drop_na(sex, agekdbrn) |>
  group_by(sex) |>
  summarise(avg = mean(agekdbrn))

```

##  `select()`, `filter()`, and `drop_na()` 

Use `select()` to pick specific columns from your dataset.  
Use `filter()` to keep rows that meet a condition.  
Use `drop_na()` to remove rows with missing (NA) values.  

<br>

. . .
 
```{r}
#| label: pipe01
#| code-line-numbers: "1|2|3|4"
#| eval: false

gss_all |>  # <1>
  select(year, sex, agekdbrn) |> # <2>
  filter(year == 2022) |> # <3>
  drop_na(sex, agekdbrn) |> # <4>

```

1.  Start with the `gss_all` data frame:
2.  Keep only the variables in the dataset that we need.
3.  Keep only the respondents from the 2022 survey
4.  Remove any observations with missing data for our key variables


##  `group_by()` and  `summarize()` 

Use `group_by()` to organize your data into groups based on one or more variables.  
Use `summarize()` to compute statistics like total, mean, or median for each group.  
<br>

. . .


```{r}
#| label: pipe02
#| code-line-numbers: "5|6"
#| output-location: column

gss_all |>  # <1>
  select(year, sex, agekdbrn) |> # <2>
  filter(year == 2022) |> # <3>
  drop_na(sex, agekdbrn) |> # <4>
  group_by(sex) |> # <5>
  summarise(freq = n()) # <6>

```

1.  Start with the `gss_all` data frame:
2.  Keep only the variables in the dataset that we need.
3.  Keep only the respondents from the 2022 survey
4.  Remove any observations with missing data for our key variables
5.  Do the next steps separately for each group in the variable
6.  Creates a new data frame with one row for each combination of grouping variables


## `dplyr()` in action

::: task
Compare the average and median age at first childbirth for U.S. men and women in 2022.
:::

. . .

```{r}
#| label: pipe03
#| code-line-numbers: "6-7|8|9"
#| output-location: column-fragment

gss_all |>  
  select(year, sex, agekdbrn) |> 
  filter(year == 2022) |> 
  drop_na(sex, agekdbrn) |> 
  group_by(sex) |> # <5>
  summarise(
    freq = n(),
    avg = mean(agekdbrn),
    med = median(agekdbrn)
    ) 

```


## `mutate()` in action 

Use `mutate()` to add new columns or change existing ones.  

. . .

::: task
What proportion of new parents were teenagers (e.g., under 18 years old)?
:::


```{r}
#| label: mutate00
#| code-line-numbers: "1-4|5|6"
#| output-location: fragment

gss_all |>  # <1>
  select(year, agekdbrn) |> # <1>
  filter(year == 2022) |> # <1>
  drop_na(agekdbrn) |> # <1>
  mutate(teen_parent = (agekdbrn < 18) * 1) |> # <2>
  summarise(proportion = mean(teen_parent)) # <3>
```

1.  Manipulate the dataset to variables and observations of interest
2.  Returns a logical vector (TRUE, FALSE, or NA) and * 1 converts it to numeric
3.  Mean of the variable, which equals the proportion of `1s`


## `mutate()` in action {.smaller}

Use `case_when()` inside `mutate()` to create values based on conditions.  

. . .

::: task
What proportion of new parents had their first child as teenagers, in their 20s, 30s, or after age 40?
:::

```{r}
#| label: mutate01
#| code-line-numbers: "1|5|6-9|10-11|13"
#| output-location: column-fragment

my_data <- gss_all |>  
  select(year, agekdbrn) |> 
  filter(year == 2022) |> 
  drop_na(agekdbrn)  |>
  mutate(age_groups = case_when(
    agekdbrn < 18 ~ "<18",
    agekdbrn >= 18 & agekdbrn <= 29 ~ "18–29",
    agekdbrn >= 30 & agekdbrn <= 39 ~ "30–39",
    agekdbrn >= 40 ~ "40+",
    TRUE ~ NA_character_)
    ) 

freq(my_data$age_groups, report.nas = FALSE, headings = FALSE)
```


## Assignment operators 

::: {.callout-tip icon="false"}
## {{< fa hat-wizard >}} Heads Up!

Overwriting datasets and variables can be intentional or unintentional.
:::

<br>

Let's make a tiny data frame to use as an example:

```{r}
#| label: tibble

df <- tibble(x = c(1, 2, 3, 4, 5), y = c("a", "a", "b", "c", "c"))
df
```




## Assignment operators (cont)

Suppose you run the following and **then** you inspect `df`.

::: {style="font-family: 'Shadows Into Light'"}
Will the `x` variable have values [1, 2, 3, 4, 5]{style="color: #18bc9c"} or [2, 4, 6, 8, 10]{style="color: #F39C12"}?
:::

<br>

::: columns
::: column
```{r}
#| label: no-assignment

df |>
  mutate(x = x * 2)
```
:::

::: {.column .fragment}
```{r}
#| label: no-assignment-results
df
```
:::
:::

. . .

> **Do something and show me**


## Assignment operators (cont)

Suppose you run the following and **then** you inspect `df`.

::: {style="font-family: 'Shadows Into Light'"}
Will the `x` variable have values [1, 2, 3, 4, 5]{style="color: #18bc9c"} or [2, 4, 6, 8, 10]{style="color: #F39C12"}?
:::

<br>

::: columns
::: column
```{r}
#| label: assignment

df <- df |>
  mutate(x = x * 2)
```
:::

::: {.column .fragment}
```{r}
#| label: assignment-results
#| code-line-numbers: "1"

df
```
:::
:::

. . .

> **Do something, save result, overwriting original**


## Assignment: Overwriting {.smaller}

::: columns
::: {.column width="50%"}
> **Do something, save result, overwriting original**

```{r}
#| label: overwrite
#| code-line-numbers: "5-6|7"
#| output-location: fragment

df <- tibble(
  x = c(1, 2, 3, 4, 5), 
  y = c("a", "a", "b", "c", "c")
)
df <- df |>
  mutate(x = x * 2)
df
```
:::

::: {.column .fragment width="50%"}
> **Do something, save result, *not* overwriting original**

```{r}
#| label: no-overwrite
#| code-line-numbers: "5-6|7"
#| output-location: fragment

df <- tibble(
  x = c(1, 2, 3, 4, 5), 
  y = c("a", "a", "b", "c", "c")
)
df_new <- df |>
  mutate(x = x * 2)
df_new
```
:::
:::

## Assignment: Overwriting (cont) {.smaller}

::: columns
::: {.column width="50%"}
> **Do something, save result, overwriting original [when you shouldn't]{style="color: #E74C3C"}**

```{r}
#| code-line-numbers: "5-8"

df <- tibble(
  x = c(1, 2, 3, 4, 5), 
  y = c("a", "a", "b", "c", "c")
)
df <- df |>
  group_by(y) |>
  summarize(mean_x = mean(x))
df
```
:::

::: {.column .fragment width="50%"}
> **Do something, save result, [not]{style="color: #18BC9C"} overwriting original [when you shouldn't]{style="color: #18BC9C"}**

```{r}
#| code-line-numbers: "5-8"

df <- tibble(
  x = c(1, 2, 3, 4, 5), 
  y = c("a", "a", "b", "c", "c")
)
df_summary <- df |>
  group_by(y) |>
  summarize(mean_x = mean(x))
df_summary
```
:::
:::


## Assignment: Overwriting (cont) {.smaller}

::: columns
::: {.column width="50%"}
> **Do something, save result, overwriting original**\
> [***data frame***]{style="color: #18BC9C"}

```{r}
#| code-line-numbers: "5-6"

df <- tibble(
  x = c(1, 2, 3, 4, 5), 
  y = c("a", "a", "b", "c", "c")
)
df <- df |>
  mutate(z = x + 2)
df
```
:::

::: {.column .fragment width="50%"}
> **Do something, save result, overwriting original**\
> [***column***]{style="color: #E74C3C"}

```{r}
#| code-line-numbers: "5-6"

df <- tibble(
  x = c(1, 2, 3, 4, 5), 
  y = c("a", "a", "b", "c", "c")
)
df <- df |>
  mutate(x = x + 2)
df
```
:::
:::

## Assignment: Overwriting (cont) {.smaller}

::::: columns
::: {.column width="50%"}
> **Do something, save result, not overwriting original.**

```{r}
#| label: mutateA
#| code-line-numbers: "1"
#| eval: false

my_data <- gss_all |>  
  select(year, agekdbrn) |> 
  filter(year == 2022) |> 
  drop_na(agekdbrn)  |>
  mutate(age_groups = case_when(
    agekdbrn < 18 ~ "<18",
    agekdbrn >= 18 & agekdbrn <= 29 ~ "18–29",
    agekdbrn >= 30 & agekdbrn <= 39 ~ "30–39",
    agekdbrn >= 40 ~ "40+",
    TRUE ~ NA_character_)) 

freq(my_data$age_groups, report.nas = FALSE, headings = FALSE)
```
:::

::: {.column .fragment width="50%"}
> **Do something and show me**

```{r}
#| label: mutateB
#| code-line-numbers: "11-14"
#| eval: false

gss_all |>  
  select(year, agekdbrn) |> 
  filter(year == 2022) |> 
  drop_na(agekdbrn)  |>
  mutate(age_groups = case_when(
    agekdbrn < 18 ~ "<18",
    agekdbrn >= 18 & agekdbrn <= 29 ~ "18–29",
    agekdbrn >= 30 & agekdbrn <= 39 ~ "30–39",
    agekdbrn >= 40 ~ "40+",
    TRUE ~ NA_character_)) |>
  group_by(age_groups) |>  
  summarise(
    count = n(),
    proportion = round(count / sum(count), 3)
  )

```
:::
:::::

# Summary Statistics {.theme-section}

## Mode

Remember, the **mode** is the category with the greatest frequency (or
the largest percentage). Let's find it for the `premarsx` variable.

<br>

```{r}
#| eval: false ## temporary!!!
#| label: mode

freq(gss_all$premarsx, report.nas = FALSE) 

```

## Median

We can use the same table we generated before to identify the
**median**. This time, let's use [dplyr grammar]() to produce the same
table

Remember, use the cumulative percentage to locate the **50th
percentile**.

```{r}
#| eval: false ## temporary!!!

#| label: median
#| code-line-numbers: "1|2|3"

gss_all |> # <1>
  freq(premarsx, report.nas = FALSE) |> # <2>
  tb() # <3> 

```

1.  Use `dplyr` grammar, starting with the name of the df and a pipe
2.  Use the `freq()` function as usual
3.  Add the `tb()` function to turn the table into a `tibble`

## Mean

```{r}
#| eval: false ## temporary!!!

#| label: freq-numeric

freq(gss_all$hrs1, report.nas = FALSE, headings = FALSE) 

```

::: aside
Question 1a. If working, full or part time: how many hours did you work
last week, at all jobs?
:::

## Mean

```{r}
#| eval: false ## temporary!!!

#| label: mean-median

mean(gss_all$hrs1, na.rm=TRUE) # <1>
median(gss_all$hrs1, na.rm=TRUE)

```

1.  `na.rm` is a logical evaluating to TRUE or FALSE indicating whether
    NA values should be stripped before the computation proceeds.

::: aside
Question 1a. If working, full or part time: how many hours did you work
last week, at all jobs?
:::

## `summary()`

<br>

```{r}
#| eval: false ## temporary!!!

#| label: summary

summary(gss_all$hrs1)

```

## Variability: `descr()`

<br>

```{r}
#| eval: false ## temporary!!!

#| label: descr01

descr(gss_all$hrs1)

```

## Variability: `descr()`

```{r}
#| label: descr02
#| code-line-numbers: "1|2|3|4"

gss_all |> 
  descr(hrs1,
        stats = "common") |> # <1>
  tb() 

```

## Variability `descr()`

```{r}
#| label: descr03
#| code-line-numbers: "1|2|3|4-5|6"

gss_all |> 
  filter(!is.na(sex))  |>
  group_by(sex) |>
  descr(hrs1,
        stats = "common") |> # <1>
  tb() 

```

1.  Which stats to produce. Either “all” (default), “fivenum”, “common”
    (see Details), or a selection. See ?descr
    
    
# Think Like a Statistician {background-image="images/lightbulb-solid.png" background-size="80px" background-repeat="repeat"}

Political polarization is high in the U.S. today and attitudes about
gender and family behavior have been heavily debated.

Using the most recent survey, do more liberals than conservatives think
sex before marriage is '*not wrong at all*'?

<br>

[{{< fa regular lightbulb >}} **How do we find
out?**]{style="color: #F39C12; font-size: 175%; font-family: 'Shadows Into Light'"}

## Variable Descriptions

Let's familiarize ourselves with the `premarsx` and `polviews`
variables.

<br>

In the console, type `?premarsx` and hit enter. The Help pane will show
you the question text, response options and values.

<br>

Now, do the same for `polviews`.

## Change variables

```{r}
#| label: think01
#| code-line-numbers: "1|2|3-5|6|7-8|9"

gss_all <- gss_all |> # <1>
 mutate(pol3cat = case_when( # <2>
   polviews >= 1 & polviews <= 3 ~ "Liberal", # <2>
   polviews == 4 ~ "Moderate", # <2>
   polviews >= 5 & polviews <= 7 ~ "Conservative", # <2>
   TRUE ~ NA_character_), # <3>
  pol3cat = factor(pol3cat, # <4>
                 levels = c("Liberal", "Moderate", "Conservative")) # <4>
  )

```

1.  Save over the dataset with an added variable.
2.  Creates a new variable by assigning labels based on values of
    `polviews`
3.  Set everything else to "missing"
4.  Convert the variable to factor with specified level order

##  {.smaller data-menu-title="Your Data Take" background-image="images/lightbulb-solid.png" background-size="80px" background-repeat="repeat"}

::: {style="color: #f39c12; font-size: 250%; font-family: 'Shadows Into Light'"}
{{< fa regular lightbulb >}} **What's your conclusion to our initial
research question?**
:::

<br>

\% who think sex relations before marriage is \_\_\_\_\_\_\_\_\_\_, by
political views

```{r}
#| label: crosstab-02-output
#| echo: false

ctable(gss_all$premarsx, gss_all$sex, prop = "c", format = "p", useNA = "no") 
```
    
    