---
title: "Simple Quarto Demo"
author: "Joanna"
format: html
---

# Introduction

This document demonstrates some basic R functionality using Quarto.

## Create Variables

```{r}
#| label: objects
name <- "Joanna"
x <- 1:10
y <- x^2
```

## Display Output

The following code prints a greeting.

```{r}
print(paste("Hello,", name))
```

## Create a Data Frame

```{r}
demo_data <- data.frame(
  Number = x,
  Square = y
)

demo_data
```

Notice that Quarto automatically displays the data frame as a table in
the rendered document.

## Basic Calculations

```{r}
sum(y)
mean(x)
```

## Create a Plot

```{r}
plot(
  x, y,
  type = "b",
  col = "orange",
  pch = 19,
  main = "Squares of Numbers",
  xlab = "Number",
  ylab = "Square"
)
```

The plot appears directly in the rendered document.

## Getting Help in R

When working interactively in RStudio, try these commands in the
**Console**:

``` r
?mean
help(plot)
```

These commands open the corresponding help pages in the **Help** pane.

## Try These in the Console

The following examples are meant to be typed directly into the Console
rather than included in the document:

``` r
2 + 2
sqrt(25)
mean(x)
```

## What This Demonstrates

- **Script/Source Pane**: The Quarto document itself
- **Console Pane**: Running code and trying commands interactively
- **Environment Pane**: Objects such as `name`, `x`, `y`, and
  `demo_data`
- **Plots Pane**: Generated plots
- **Help Pane**: Documentation opened with `?mean`
- **Viewer Pane**: Rendered Quarto HTML document
