---
title: "Code-Along 01"
format: html
---

## Packages

We'll use the following packages:

-   `here()` (relative file paths)
-   `tidyverse()` (data wrangling)
-   `gssr()` (U.S. General Social Survey data)
-   `gssrdoc()` (GSS documentation)

## Install `here()` and `tidyverse()`

Let's first install the two packages that are available on CRAN. 

Copy and paste the following code into your Console pane. Then hit enter.

``` {r}
#| eval: false
#| message: false

install.packages("here")
```

Then, do the same to install the tidyverse package.

``` {r}
#| eval: false
#| message: false

install.packages("tidyverse")
```


## Install `gssr()` and `gssrdoc()`

```{r}
# Install 'gssr' from 'ropensci' universe
install.packages('gssr', repos =
  c('https://kjhealy.r-universe.dev', 'https://cloud.r-project.org'))

# Also recommended: install 'gssrdoc' as well
install.packages('gssrdoc', repos =
  c('https://kjhealy.r-universe.dev', 'https://cloud.r-project.org'))
```

## Load the packages

```{r}
library(here)
library(tidyverse)
library(gssr)
library(gssrdoc)
```
