---
title: "Code-along 06"
author: "FirstName LastName"
editor: visual
execute:
  error: true
format: pdf # or replace 'pdf' with 'typst'
---

# Setup

## Packages

Install the [`gtsummary()`](https://www.danieldsjoberg.com/gtsummary/)
package, available on CRAN.

```{r}
#| eval: false
#| label: install-Rmisc

install.packages("gtsummary")
```

Load the standard packages and our new `gtsummary()` package.

```{r}
#| message: false
#| label: load

library(tidyverse)
library(haven) # not core tidyverse
library(gssr)
library(gssrdoc) # load the GSS codebook
library(summarytools)
library(gtsummary) # load the new package
```

## Load your data

Get the data from the 2022 survey

```{r}
#| label: gssdata

# ADD YOUR CODE HERE

```

Let's make a df with only the (pretty) categorical and continuous
variables we'll analyze.

```{r}
#| label: my-data

# Categorical Variables
my_cat <- gss22 |>
  select(id, premarsx, fefam, postlife, sex, polviews) |>
  zap_missing() |>
  as_factor() |>
  droplevels()

# Continuous Variables
my_con <- gss22 |>
  select(id, age, educ, lifenow, agekdbrn) |>
  mutate(
    age = as.numeric(age),
    educ = as.numeric(educ),
    lifenow = as.numeric(lifenow),
    agekdbrn = as.numeric(agekdbrn))

# Combine the two dataframes
my_data <- left_join(my_cat, my_con, by = "id")
```

# $chi^2$ Test

## Review: Pretty cross-tab

Use `summarytools::ctable`

```{r}
#| label: ctable

ctable(my_data$fefam, my_data$sex, 
  prop = "c", 
  format = "p", 
  useNA = "no"
) 
```

## Pretty cross-tab (new)

Now that we know how to use the `dplyr |>`!

```{r}
#| label: tbl-cross

my_data |>
  tbl_cross(
    row = fefam,
    col = sex,
    percent = "column",
    missing = "no")
```

## Pretty cross-tab with `p-value`

```{r}
#| label: tbl-cross-p

my_data |>
  tbl_cross(
    row = fefam,
    col = sex,
    percent = "column",
    missing = "no") |>
  add_p(source_note = TRUE)
```

## Caution: size matters

```{r}
#| label: sample-size

my_data |>
  filter(age < 21) |>
  tbl_cross(
    row = fefam,
    col = sex,
    percent = "column",
    missing = "no") |>
  add_p(source_note = TRUE)
```

## Review: `t.test()` with proportions

```{r}
#| label: t.test

# Use logic to recode factor as mean (yes = 1; no: 2 = 0)
my_data$ghosts <- ifelse(my_data$postlife == "yes", 1, 0)

# t.test
t.test(my_data$ghosts ~ my_data$sex, alternative = "two.sided")

```

## `t.test()` & [$chi^2$]{style="color: #18BC9C"}

```{r}
#| message: false
#| label: postlife

my_data |>
  tbl_cross(
    row = ghosts,
    col = sex,
    percent = "column",
    missing = "no") |>
  add_p(source_note = TRUE)

```

# Pearson’s Correlation Coefficient (R)

R is the correlation between two interval-ratio variables.

## `cor.test()` in action

```{r}
#| label: cor-educ01

cor.test(
  ~ educ + agekdbrn,
  data = my_data,
  method = "pearson",
  na.action = na.omit
)

```

## Tidy `cor.test()`

```{r}
#| label: cor-educ02

# Save results
my_test <- cor.test(
  ~ educ + agekdbrn,
  data = my_data,
  method = "pearson",
  na.action = na.omit
)
```

```{r}
#| label: cor-educ03

# Create a tibble with key stats
cor_summary <- tibble(
  correlation = my_test$estimate,
  p_value     = my_test$p.value,
  conf_low    = my_test$conf.int[1],
  conf_high   = my_test$conf.int[2]
)

cor_summary
```

# Think Like a Statistician

**Are beliefs about premarital sex (`premarsx`) dependent on political
identity (`polviews`)?**

```{r}
#| label: tbl-cross-premarsx

# YOUR CODE HERE

```

**Is there a gender difference (`sex`) in political identity
(`polviews`)?**

```{r}
#| label: tbl-cross-polviews

# YOUR CODE HERE

```

**Describe the relationship between `age`and life satisfaction
(`lifenow`).**

```{r}
#| label: cor-lifenow

# YOUR CODE HERE

```
