---
title: "Code-along 01"
author: "FirstName LastName"
editor: visual
execute:
  error: true
format: pdf # or replace 'pdf' with 'typst'
---

## Packages

We'll use the following packages:

-   `tidyverse()` (data wrangling)
-   `gssr()` (U.S. General Social Survey data)
-   `gssrdoc()` (GSS documentation)

## Load the packages

```{r}
#| label: load

library(tidyverse)
library(gssr)
library(gssrdoc)
```

## Environment

```{r}
#| label: enviro

# software documentation
sessionInfo()

```

## Meet your data

We're going to use data from the [U.S. General Social Survey
(GSS)](https://gss.norc.org/us/en/gss/about-the-gss.html).

Let's load your data.

```{r, eval=FALSE}
#| label: gss24

# Get the data only for the 2024 survey respondents
gss24 <- gss_get_yr(2024)

# look at the dataframe
gss24

```

A “tibble” is another name for “tidy dataset,” meaning that the data is
organized in structured, clear rows and columns. “(75,699 × 6,867)”
means the dataset contains 75,699 rows and 6,867 columns. In social
sciences, rows are commonly referred to as “observations” and columns as
“variables.” In our case, there are 75,699 observations (e.g.,
respondents) and 6,867 variables.

You can also load the GSS data for a specific survey year.

With your mouse, go to the environment panel (upper-right) and click on
the `gss24` object. It pops up and you can browse through it.

This is often a good idea to get a first feel for the data, but only if
your dataset is relatively small.

## Variables

The [GSS
documentation](https://gss.norc.org/us/en/gss/get-documentation.html) is
also available online in .pdf form.

The .pdfs will be useful for general overviews.

For specific variable information, it will be helpful to use the
documentation you'll load into RStudio.

For information about a specific GSS variable, type `?varname` at the
console.

In the output pane, the Help tab will show the variable documentation.

Example:

```{r, eval=FALSE}
#| label: meovrwrk-help
#| eval: false 
 
# copy/paste this code in the Console; don't run in Quarto
?meovrwrk
```

Let's look at a frequency table of this variable.

(NOTE: The variable names are case sensitive. In this dataset, all
variables are lowercase.)

```{r}
#| label: meovrwrk-table-gss_all

table(gss_all$meovrwrk)

```

2436 respondents were coded as 1 on this variable. What does that mean?
(Look at the help page.)

**Change the code to show the variable `fefam`.**

Then, add text to your code-along that interprets the results for the
`2` value.

```{r}
#| label: fefam-table-gss24

# Replace meovrwrk with fefam
table(gss24$meovrwrk)
```

## Quarto: Render

Finally, let's `render` your code-along-01 and view the results.
