---
title: "Code-along 05"
author: "FirstName LastName"
editor: visual
execute:
  error: true
format: pdf # or replace 'pdf' with 'typst'
---

# Setup

## Packages

Install the
[`Rmisc()`](https://cran.r-project.org/web/packages/Rmisc/refman/Rmisc.html)
package, available on CRAN.

```{r}
#| eval: false
#| label: install-Rmisc

install.packages("Rmisc")
```

Load the standard packages and our new `Rmisc()` package.

```{r}
#| message: false
#| label: load

library(tidyverse)
library(haven) # not core tidyverse
library(gssr)
library(gssrdoc) # load the GSS codebook
library(summarytools)
library(Rmisc) # load the new package
```

## Load your data

Get the data from the 2022 survey

```{r}
#| label: gssdata

# ADD YOUR CODE HERE

```

Create a dataset named `my_data` with the variables `wwwhr`, `postlife`, `childs`, &
`sex` with no missing data

```{r}
#| message: false
#| label: my_data

# ADD YOUR CODE HERE

```

# Confidence Intervals

## `summarise()` with equation

```{r}
#| label: summarise02

wwwhr_sum <- my_data |>
  summarise(
    n = n(),
    mean_wwwhr = mean(wwwhr),
    sd_wwwhr = sd(wwwhr),
    se_wwwhr = sd_wwwhr / sqrt(n),
    ci_lower = mean_wwwhr - (1.96 * se_wwwhr),
    ci_upper = mean_wwwhr + (1.96 * se_wwwhr)
    )

```

## Code Meets Question

What can we infer about the population mean based on this confidence
interval?

```{r}
#| label: wwwhr_sum-CMQ

wwwhr_sum

```

## `summarise()` + `CI()`

```{r}
#| label: summarise-ci95

wwwhr_ci <- my_data |>
  summarise(
    n = n(),
    mean_wwwhr = mean(wwwhr),
    sd_wwwhr = sd(wwwhr),
    lowCI_wwwhr = Rmisc::CI(wwwhr)['lower'],
    hiCI_wwwhr  = Rmisc::CI(wwwhr)['upper']
    )

wwwhr_ci
```

## `summarise()` + `CI()`

```{r}
#| label: summarise-ci99

wwwhr_ci99 <- my_data |>
  summarise(
    n = n(),
    mean_wwwhr = mean(wwwhr),
    sd_wwwhr = sd(wwwhr),
    lowCI_wwwhr = CI(wwwhr, ci = 0.99)['lower'],
    hiCI_wwwhr  = CI(wwwhr, ci = 0.99)['upper']
    )

wwwhr_ci99
```

::: task
{{< fa lightbulb >}} **How does a 95% confidence interval differ from a
99% confidence interval in terms of width and certainty?**
:::

## CIs (wrong) with proportions

```{r}
#| label: proportions01

my_data |>
  summarise(
    n = n(),
    mean_ghosts = mean(postlife),
    sd_ghosts = sd(postlife),
    lowCI_ghosts = CI(postlife, ci = 0.99)['lower'],
    hiCI_ghosts  = CI(postlife, ci = 0.99)['upper']
    )

```

## CIs (correct) with proportions

```{r}
#| label: proportions02

# Use logic to recode (yes: 1 = 1; no: 2 = 0)
my_data$ghosts <- ifelse(my_data$postlife == 1, 1, 0)
  
my_data |>
  summarise(
    n = n(),
    mean_ghosts = mean(ghosts),
    sd_ghosts = sd(ghosts),
    lowCI_ghosts = CI(ghosts, ci = 0.99)['lower'],
    hiCI_ghosts  = CI(ghosts, ci = 0.99)['upper']
    )

```

## Comparing CIs


```{r}
#| label: proportions03

my_data$sex <- zap_missing(my_data$sex) 
my_data$sex <- as_factor(my_data$sex)   
my_data$sex <- droplevels(my_data$sex)  

df_ghosts <- my_data |>
  group_by(sex) |>
  summarise(
    n = n(),
    mean = mean(ghosts),
    sd = sd(ghosts),
    lower = CI(ghosts, ci = 0.95)['lower'],
    upper  = CI(ghosts, ci = 0.95)['upper']
    )

df_ghosts

```


```{r}
#| label: wwwhr01
#| code-line-numbers: "1-2|3-9"

df_wwwhr <- my_data |>
  group_by(sex) |>
  summarise(
    n = n(),
    mean = mean(wwwhr),
    sd = sd(wwwhr),
    lower = CI(wwwhr, ci = 0.95)['lower'],
    upper  = CI(wwwhr, ci = 0.95)['upper']
    )

df_wwwhr

```


# Hypothesis Testing 

## One sample `t.test()` for means

```{r}
#| label: ttest01

t.test(my_data$childs)

```

> **H0 = 0 by default**

## One sample `t.test()` for means

```{r}
#| label: ttest02

t.test(my_data$childs, mu=2, alternative="less") # Ho: mu=2

```

> **H0 = 2; one-tail test**


## two sample `t.test()` in action

Are the proportions of men and women who believe in ghosts statistically significantly different?

```{r}
#| label: ttest03

t.test(my_data$ghosts ~ my_data$sex, alternative = "two.sided")

```

 Assuming an α level of .05, what do you conclude?

## Aside: paired `t.test()` 

**Independent group t-test**\
Do two subsamples have the same mean?

```{r}
#| label: ttest-samples

t.test(my_data$ghosts ~ my_data$sex, alternative = "two.sided")

```


**Paired t-test**\
Do two variables have the same mean?

```{r}
#| message: false
#| label: ttest-paired
#| eval: false

t.test(Pair(hrs1_2018, hrs1_2020) ~ 1, data = gss_panel)

```

# Think Like a Statistician 

(*Use the 2022 GSS dataset.*)

**Do women have more years of education (`educ`) than men, on average?**

```{r}
#| label: educ01

# YOUR CODE HERE

```

**Are the proportions of men & women who support abortion for any reason (`abany`) statistically different?**

```{r}
#| label: abany01

# YOUR CODE HERE

```

 **Do fewer men than women support gun laws (`gunlaw`)?**

```{r}
#| label: gunlaw01

# YOUR CODE HERE

```

