---
title: "Code-along 01"
author: "FirstName LastName"
editor: visual
execute:
  error: true
format: pdf # or replace 'pdf' with 'typst'
---

# Packages

## Install Packages

**Click the green arrow icon to run the following code chunks.**

These chunks will not run (on purpose!) when rendered because
`eval: false`.

**Install the `tidyverse()` package.**

```{r}
#| eval: false
#| message: false
#| label: install-tidyverse

install.packages("tidyverse")
```

**Install the `tinytex()` package.** This package is needed to Render
Quarto documents as .pdf.

```{r}
#| eval: false
#| message: false
#| label: install-tinytex01

install.packages('tinytex')
```

**Then, use `tinytex` package to install a part of the package.**

```{r}
#| eval: false
#| message: false
#| label: install-tinytex02

tinytex::install_tinytex()
```

**Install the `gssr()` package, NOT on CRAN.**

```{r}
#| eval: false
#| message: false
#| label: install-gssr

# Install 'gssr' from 'ropensci' universe
install.packages('gssr', repos =
  c('https://kjhealy.r-universe.dev', 'https://cloud.r-project.org'))
```

**Install the `gssrdoc()` package, NOT on CRAN.**

```{r}
#| eval: false
#| message: false
#| label: install-gssrdoc

# Install 'gssrdoc' from 'ropensci' universe
install.packages('gssrdoc', repos =
  c('https://kjhealy.r-universe.dev', 'https://cloud.r-project.org'))
```

## Load the packages

We'll use the following packages:

-   `tidyverse()` (data wrangling)
-   `gssr()` (U.S. General Social Survey data)
-   `gssrdoc()` (GSS documentation)

```{r}
#| label: load

library(tidyverse)
library(gssr)
library(gssrdoc)
```

## Environment

```{r}
#| label: enviro

# software documentation
sessionInfo()

```

## Meet your data

We're going to use data from the [U.S. General Social Survey
(GSS)](https://gss.norc.org/us/en/gss/about-the-gss.html).

Let's load your data.

```{r}
#| label: gss24

# Get the data only for the 2024 survey respondents
gss24 <- gss_get_yr(2024)

# look at the dataframe
gss24

```

A “tibble” is another name for “tidy dataset,” meaning that the data is
organized in structured, clear rows and columns. “(3,309 × 639)” means
the dataset contains 3,309 rows and 639 columns. In social sciences,
rows are commonly referred to as “observations” and columns as
“variables.” In our case, there are 3,309 observations (e.g.,
respondents) and 639 variables.

With your mouse, go to the environment panel (upper-right) and click on
the `gss24` object. It pops up and you can browse through it.

This is often a good idea to get a first feel for the data, but only if
your dataset is relatively small.

## Variables

The [GSS
documentation](https://gss.norc.org/us/en/gss/get-documentation.html) is
also available online in .pdf form.

The .pdfs will be useful for general overviews.

For specific variable information, it will be helpful to use the
documentation you'll load into RStudio.

For information about a specific GSS variable, type `?varname` at the
console.

In the output pane, the Help tab will show the variable documentation.

Example:

```{r, eval=FALSE}
#| label: meovrwrk-help
#| eval: false 
 
# copy/paste this code in the Console; don't run in Quarto
?meovrwrk
```

Let's look at a frequency table of this variable.

(NOTE: The variable names are case sensitive. In this dataset, all
variables are lowercase.)

```{r}
#| label: meovrwrk-table-gss24

table(gss24$meovrwrk)

```

195 respondents were coded as 1 on this variable. What does that mean?
(Look at the help page.)

**Change the code to show the variable `fefam`.**

*Add text here that interprets the results for the `2` value for this
variable.*

```{r}
#| label: fefam-table-gss24

# Replace meovrwrk with fefam
table(gss24$meovrwrk)
```

## Quarto: Render

Finally, `Render` your code-along-01 and view the results.
