---
title: "Code-along 09"
author: "FirstName LastName"
editor: visual
execute:
  error: true
format: pdf # or replace 'pdf' with 'typst'
---

# Setup

## Packages

Load the standard packages and new package `modelbased`.

```{r}
#| message: false
#| label: load

# install.packages("marginaleffects") # needed for modelbased
# install.packages("modelbased")

library(tidyverse) 
library(haven) # not core tidyverse
library(gssr)
library(gssrdoc)
library(summarytools)
library(gtsummary) 
library(moderndive)
library(modelbased)
```

## Load your data

```{r}
#| label: gssdata

# Load all gss
data(gss_all)

```

Make a df with only the (pretty) categorical and continuous variables
we'll analyze.

```{r}
#| label: my-data

# Categorical Variables
my_cat <- gss_all |>
  select(id, year, polviews, sex, marital) |>
  zap_missing() |>
  mutate(
    polviews = as_factor(polviews),
    sex = as_factor(sex),
    marital = as_factor(marital)) |> 
  droplevels()

# Continuous Variables
my_con <- gss_all |>
  select(id, year, tvhours, age, educ, hrs1, realrinc, childs) |>
  mutate(
    age = as.numeric(age),
    educ = as.numeric(educ),
    hrs1 = as.numeric(hrs1),
    realrinc = as.numeric(realrinc),
    childs = as.numeric(childs))

# Combine the two dataframes
my_data <- full_join(my_cat, my_con, by = c("year", "id")) # match by 2 vars

```

# Two continuous variables

> ***Have the average number of children changed over time?***

```{r}
#| label: plot-01
#| message: false

# Create scatterplot with a regression line
my_data |>
  group_by(year) |>
  summarize(avg = mean(childs, na.rm = TRUE)) |>
  ggplot(aes(x = year, y = avg)) +
  geom_point(color = "#18BC9C", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "#3498DB") +
  labs(
    title = "Scatterplot of Average Number of Children",
    x = "Year",
    y = "Average number of children") +
  theme_minimal()
```

**Pearson's Correlation (R)**

```{r}
#| label: cor.test

cor.test(
  ~ childs + year,
  data = my_data,
  method = "pearson",
  na.action = na.omit)
```

**Simple Regression**

```{r}
#| label: regress-01

model_01 <- lm(formula = childs ~ year, data = my_data)
get_regression_table(model_01)

```

**PRE Measure**

Square R

```{r}
#| label: r-squared

my_test <-  cor.test(
  ~ childs + year,
  data = my_data,
  method = "pearson",
  na.action = na.omit
)

my_test$estimate^2

```

Get $r^2$

```{r}
#| label: get_regression_summaries

model_01 <- lm(formula = childs ~ year, data = my_data)
get_regression_summaries(model_01)
```

> ***Are the trends explained by increases in education?***

```{r}
#| label: plot-02
#| message: false

# Create scatterplot with regression line
my_data |>
  group_by(year) |>
  summarize(avg = mean(educ, na.rm = TRUE)) |>
  ggplot(aes(x = year, y = avg)) +
  geom_point(color = "#18BC9C", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "#3498DB") +
  labs(
    title = "Scatterplot of Average Years of Education",
    x = "Year",
    y = "Average years of education"
  ) +
  theme_minimal()
```

```{r}
#| label: plot-03
#| message: false

# Create scatterplot with regression line
my_data |>
  group_by(educ) |>
  summarize(avg = mean(childs, na.rm = TRUE)) |>
  ggplot(aes(x = educ, y = avg)) +
  geom_point(color = "#18BC9C", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "#3498DB") +
  labs(
    title = "Scatterplot of Average Number of Children",
    x = "Average years of education",
    y = "Average number of children"
  ) +
  theme_minimal()
```

## Multiple Regression

```{r}
#| label: regress-02

model_02 <- lm(formula = childs ~ year + educ, data = my_data)
get_regression_table(model_02)

```

**PRE Measure**

```{r}
#| label: summaries-02

get_regression_summaries(model_02)

```

## `tbl_regression()`

```{r}
#| label: tbl_regression

model_02 |>
  tbl_regression(intercept = TRUE)  

```

```{r}
#| label: tbl_regression-partial

model_02 |>
  tbl_regression(
    intercept = TRUE) |>
  modify_column_hide(conf.low) |>
  modify_column_unhide(column = std.error) 

```

**OLS Regression predicting number of children by survey year and years
of education**

```{r}
#| label: tbl_regression-full

model_02 |>
  tbl_regression(
    intercept = TRUE) |>
  modify_column_hide(conf.low) |>
  modify_column_unhide(column = std.error) |>
  remove_abbreviation() |>
  add_significance_stars() |>
  add_glance_table(include = c("r.squared", "nobs"))

```

# 1 continuous variable & 1 categorical variable

> ***Are weekly hours worked related to education and gender?***

```{r}
#| label: regress-03

model <- lm(hrs1 ~ educ + sex, data = my_data)

get_regression_table(model)
```

```{r}
#| label: regress-03-table

model |>
  tbl_regression(
    intercept = TRUE) |>
  modify_column_hide(conf.low) |>
  modify_column_unhide(
    column = std.error) |>
  remove_abbreviation() |>
  add_significance_stars() |>
  add_glance_table(
    include = c("r.squared", "nobs"))

```

**OLS Regression predicting weekly work hours by years of education and
gender**

```{r}
#| label: dichotomous

model |>
  tbl_regression(
    intercept = TRUE,
    show_single_row = sex,
    label = list(sex = "Women")) |>
  modify_column_hide(conf.low) |>
  modify_column_unhide(
    column = std.error) |>
  remove_abbreviation() |>
  add_significance_stars() |>
  add_glance_table(
    include = c("r.squared", "nobs"))

```

## `estimate_means()`

R calculates the mean values of every variable in the model except sex,
and sets all of those variables equal to their means. Then (once again)
it sets female = 0 in every observation, calculates predicted
probabilities for each individual, and averages those.

Then it goes back and resets female to 1 in every observation, again
calculates predicted probabilities for each individual and outputs the
average of those.

Outputs can be interpreted as the predicted probabilities for women and
men if they were all average in all other respects.

```{r}
#| label: estimate_means_margins

# Get predicted values
preds <- estimate_means(model, by = c("sex"))

# Preview the new df
preds |>
  as_tibble() |> 
  head()
```

```{r, eval = FALSE}
#| label: estimate_means_code

# Get predicted values
preds <- estimate_means(model, by = c("educ", "sex"))

# Preview the new df
preds |>
  as_tibble() |> 
  head()
```

men with 0 years of education = $38.601 + .420_{educ}*0$\
women with 0 years of education
$38.601 + .420_{educ}*0 -6.576_{female}$\
men with 2.222 years of education $38.601 + .420_{educ}*2.222$\
women with 2.222 years of education
$38.601 + .420_{educ}*2.222-6.576_{female}$

**Visualize the predictions:**

```{r}
#| label: ggplot
#| message: false

# Get predicted values
preds <- estimate_means(model, by =c("educ", "sex"))

# Plot using ggplot2
preds |>
  ggplot(aes(x = educ, y = Mean, color = sex)) +
    geom_line(linewidth = 1.2) +
    labs(
      title = "Predicted Weekly Work Hours by Education and Gender",
      x = "Education Level",
      y = "Predicted Weekly Work Hours",
      color = " ") +
    theme_minimal() +
    theme(legend.position = "top")
```

> ***Are weekly hours worked related to education and marital status?***

```{r}
#| label: regress-04

model <- lm(hrs1 ~ educ + marital, data = my_data)
get_regression_table(model)

```

**Regression Predicting Weekly Work Hours by Years of Education and
Marital Status**

```{r}
#| label: reference

model |>
  tbl_regression(
    intercept = TRUE,
    label = list(
      marital = "Marital status (ref. = married)")) |>
  remove_row_type(type = c("reference")) |>  
  modify_column_hide(conf.low) |>
  modify_column_unhide(
    column = std.error) |>
  remove_abbreviation() |>
  add_significance_stars() |>
  add_glance_table(
    include = c("r.squared", "nobs"))

```

> ***Are daily TV hours related to education and political identity?***

```{r}
#| label: regress-05

my_data <- my_data %>%
  mutate(pol_group = case_when(
    polviews %in% c("extremely liberal", "liberal", "slightly liberal") ~ "Liberal",
    polviews == "moderate, middle of the road" ~ "Moderate",
    polviews %in% c("slightly conservative", "conservative", "extremely conservative") ~ "Conservative",
    TRUE ~ NA_character_))

model <- lm(tvhours ~ educ + pol_group, data = my_data)

get_regression_table(model)
```

## `relevel()`

```{r}
#| label: relevel

# Relevel pol_group so "Moderate" is the reference
my_data$pol_group <- relevel(factor(my_data$pol_group), 
                            ref = "Moderate")

model <- lm(tvhours ~ educ + pol_group, data = my_data)
get_regression_table(model)
```

> ***Does the relationship between TV hours and education differ by
> political identity?***

```{r}
#| label: regress-06

model <- lm(tvhours ~ educ * pol_group, data = my_data)

get_regression_table(model)
```

The coefficient for `educ` is \$-0.142 \$, meaning that for moderates
(the reference group), each additional year of education is associated
with -0.142 fewer tv hours per day.

The coefficient for `pol_group: Conservative` is $-0.1$. This is the
difference in the outcome between moderates and conservatives when
education = 0. Conservatives watch -0.1 TV hours per day fewer than
moderates when education is zero. Not statistically significantly
different.

The interaction term `educ:pol_groupConservative` is $-0.008$. It tells
us whether the relationship between education and TV hours differs
between moderates and conservatives It adjusts the slope of education
for conservatives So for conservatives, the effect of education is:

$-0.142 -0.008 = -0.15$

This means that for conservatives, each additional unit of education is
associated with only $-0.15$ fewer daily TV hours.

**Visualize predictions:**

```{r}
#| label: ggplot-interaction
#| message: false

# Get predicted values
preds <- estimate_means(model, by = c("educ", "pol_group"))

# Plot using ggplot2
preds |>
  ggplot(aes(x = educ, y = Mean, color = pol_group)) +
    geom_line(linewidth = 1.2) +
    labs(
      title = "Predicted Daily TV Hours by Education and Political Identity",
      x = "Education Level",
      y = "Predicted TV Hours",
      color = "Political Identity") +
    theme_minimal() +
    theme(legend.position = "top")
```
