---
title: "Code-along 04"
author: "FirstName LastName"
editor: visual
execute:
  error: true
format: pdf # or replace 'pdf' with 'typst'
---

# Setup

## Packages

Load the standard packages.

```{r}
#| message: false
#| label: load

library(tidyverse) 
library(haven) # not core tidyverse
library(gssr)
library(gssrdoc) # load the GSS codebook
library(summarytools)
```

## Load your data

```{r}
#| message: false
#| label: gssdata

# Get the data all survey years
data(gss_all)

```

# dplyr II

## `select()`

Use `select()` to choose columns in a dataframe.\

```{r}
#| label: select

my_data <- gss_all |> 
  select(year, agekdbrn, childs, hrs1) 

```

## `head()` & `tail()`

`head()` looks at the first few column names and **first** few rows of
your df.\
`tail()` looks at the first few column names and **last** few rows of
your df.

```{r}
#| label: head

# look at the first rows & columns of the data frame
head(my_data)

```

**Use `tail()` to look at the last few rows of `my_data`**

```{r, eval=FALSE}
#| label: tail

# ADD YOUR CODE HERE

```

**Create a dataframe with the variables `year`, `agekdbrn`, `childs`,
and `hrs1`, including only respondents from the year 2022 who have no
missing values in any of these variables.**

```{r, eval=FALSE}
#| label: select-mini-task

my_data <- gss_all |> 
  # ADD YOUR CODE HERE

head(my_data) 

```

## `mutate()`

Use `mutate()` to add new columns or change existing ones.

```{r}
#| label: mutate

# Use mutate to create a new variable: annual_hrs1
my_data <- my_data |>
  mutate(annual_hrs1 = hrs1 * 52)

head(my_data)
```

**Use `mutate()`, `mean()`, and `sd()` to create a column with Z scores
for `hrs1`**

```{r, eval=FALSE}
#| label: mutate-mini-task

my_data <- my_data |>
  # ADD YOUR CODE HERE

head(my_data) # Check your work

```

## `mutate()` & logical variables

```{r}
#| label: mutate-logical

my_data <- my_data |>
  mutate(teen_parent = (agekdbrn < 18) * 1) 
  
head(my_data)
  
```

## Code Meets Question

**What proportion of new parents were teenagers (e.g., under 18 years
old)?**

```{r}
#| label: mutate-logical-CMQ

my_data |> 
  summarise(proportion = mean(teen_parent)) 
  
```

**Use `mutate()` and `summarise()` to determine what proportion of new
parents were older than 40 years old.**

```{r, eval=FALSE}
#| label: mutate-logical-mini-task

# ADD YOUR CODE HERE
  
```

## `mutate()` with `case_when()`

Use `case_when()` inside `mutate()` to create values based on
conditions.

```{r}
#| label: mutate-case-when

my_data <-  my_data |>  
# Use mutate to create a new variable: childs_3cat
  mutate(childs_3cat = case_when(
    childs == 1 ~ "1 kid",
    childs == 2 ~ "2 kids",
    childs >= 3 ~ "3+ kids",
    TRUE ~ NA_character_)
  ) 

# Always double check your work!
table(my_data$childs, my_data$childs_3cat)
```

## Code Meets Question

**What proportion of new parents had their first child in their teens,
20s, 30s, or after age 40?**

```{r, eval=FALSE}
#| label: mutate01

my_data <-  my_data |>  
  mutate(age_groups = case_when(
    # ADD YOUR CODE HERE
  ))

my_data |>
  freq(age_groups, report.nas = FALSE, headings = FALSE)
```

# Overwriting data

## Overwriting & assignment operators

Let's make a tiny data frame to use as an example:

```{r}
#| label: tibble

df <- tibble(
        x = c(1, 2, 3, 4, 5), 
        y = c("a", "a", "b", "c", "c")
      )

df
```

> **Do something and show me**

```{r}
#| label: no-assignment

df |>
  mutate(x = x * 2)

```

> **Do something, save result, overwriting original**

```{r}
#| label: overwrite

df <- tibble(
  x = c(1, 2, 3, 4, 5), 
  y = c("a", "a", "b", "c", "c")
)

df <- df |>
  mutate(x = x * 2)

df
```

> **Do something, save result, *not* overwriting original**

```{r}
#| label: no-overwrite

df <- tibble(
  x = c(1, 2, 3, 4, 5), 
  y = c("a", "a", "b", "c", "c")
)

df_new <- df |>
  mutate(x = x * 2)

df
```

> **Do something, save result, overwriting original [when you
> shouldn't]{style="color: #E74C3C"}**

```{r}
#| label: overwrite-bad

df <- tibble(
  x = c(1, 2, 3, 4, 5), 
  y = c("a", "a", "b", "c", "c")
)
df <- df |>
  group_by(y) |>
  summarize(mean_x = mean(x)
            )

df
```

> **Do something, save result, [not]{style="color: #18BC9C"} overwriting
> original [when you shouldn't]{style="color: #18BC9C"}**

```{r}
#| label: no-overwrite-good

df <- tibble(
  x = c(1, 2, 3, 4, 5), 
  y = c("a", "a", "b", "c", "c")
  )

df_summary <- df |>
  group_by(y) |>
  summarize(mean_x = mean(x)
            )

df
```

> **Do something, save result, overwriting original** [***data
> frame***]{style="color: #18BC9C"}

```{r}
#| label: overwrite-df

df <- tibble(
  x = c(1, 2, 3, 4, 5), 
  y = c("a", "a", "b", "c", "c")
  )

df <- df |>
  mutate(z = x + 2)

df
```

> **Do something, save result, overwriting original**
> [***column***]{style="color: #f39c12"}

```{r}
#| label: overwrite-column

df <- tibble(
  x = c(1, 2, 3, 4, 5), 
  y = c("a", "a", "b", "c", "c")
  )

df <- df |>
  mutate(x = x + 2)

df
```

> **Do something and show me**

```{r}
#| label: mutate-A
#| eval: false

gss_all |>  
  select(year, agekdbrn) |> 
  filter(year == 2022) |> 
  drop_na()  |>
  mutate(age_groups = case_when(
    agekdbrn < 18 ~ "<18",
    agekdbrn >= 18 & agekdbrn <= 29 ~ "18–29",
    agekdbrn >= 30 & agekdbrn <= 39 ~ "30–39",
    agekdbrn >= 40 ~ "40+",
    TRUE ~ NA_character_)) |>
  group_by(age_groups) |>  
  summarise(
    count = n(),
    proportion = round(count / sum(count), 3)
  )

```

> **Do something, save result, not overwriting original.**

```{r}
#| label: mutate-B
#| eval: false

# Do something
gss_all <-  gss_all |>  
  mutate(age_groups = case_when(
    agekdbrn < 18 ~ "<18",
    agekdbrn >= 18 & agekdbrn <= 29 ~ "18–29",
    agekdbrn >= 30 & agekdbrn <= 39 ~ "30–39",
    agekdbrn >= 40 ~ "40+",
    TRUE ~ NA_character_)
  ) 

# Now show me
gss_all |>
  select(year, age_groups) |> 
  filter(year == 2022) |> 
  drop_na()  |>
  group_by(age_groups) |>  
  summarise(
    count = n(),
    proportion = round(count / sum(count), 3)
  )

```

## Overwriting & raw data

[Never overwrite your raw data
set!]{style="color: #E74C3C; font-size: 175%; font-family: 'Shadows Into Light'"}

Always save transformed data as a new object, which should remain
untouched for reference and reproducibility.

## `write_rds()` & `read_rds()`

`write_rds()` saves an R object to a file\
`read_rds()` loads an R object from a .rds file

```{r, eval=FALSE}
#| label: save-load

write_rds(my_data, "my_data_file.rds") 

read_rds("my_data_file.rds") 

```

# Think Like a Statistician

In 2024, how did the average amount of television watched differ between
married people and those who were never married? And how did it compare
to people who were widowed, divorced, or separated?

*How do we find out?*

```{r, eval=FALSE}
#| label: think-output

# ADD YOUR CODE HERE

```

## Your Data Take

**What's your conclusion to our research question?**
