---
title: "Code-along 08"
author: "FirstName LastName"
editor: visual
execute:
  error: true
format: pdf # or replace 'pdf' with 'typst'
---

# Setup

## Packages

Load the standard packages and install `moderndrive`

```{r}
#| message: false
#| label: load

# install.packages("moderndive")

library(here)
library(tidyverse) 
library(haven) # not core tidyverse
library(summarytools)
library(gtsummary) 
library(moderndive)
```

## Create an example dataframe

```{r}
#| label: df

# Create columns
literacy <- c(31.7, 79.3, 85.4, 41.0, 54.7, 97.9, 92.8, 67.1, 97.8, 60.4, 87.5, 99.1)
educ <- c(3.74, 10.4, 20.4, 4.46, 9.53, 78.6, 31.5, 16.2, 65.8, 10.3, 30.8, 40.1)

# Combine into a data frame
df <- data.frame(literacy, educ)

# View the data frame
print(df)

```

## Look at descriptive statistics

`moderndive::tidy_summary()` returns commonly used summary statistics in
a tidy format.

```{r}
#| label: tidy-summary

df |>
  tidy_summary()

```

Create a scatterplot.

```{r}
#| label: scatterplot

# Create scatterplot
df |>
  ggplot(aes(x = literacy, y = educ)) +
  geom_point(color = "#18BC9C", size = 3) +
  labs(
    title = "Scatterplot of Education vs Literacy",
    x = "Literacy Rate (X)",
      y = "College enrollment (Y)"
  ) +
  theme_minimal()
```

Add a regression line to the scatterplot.

```{r}
#| label: scatterplot-ols

# Create scatterplot with regression line
df |>
  ggplot(aes(x = literacy, y = educ)) +
    geom_point(color = "#18BC9C", size = 3) +
    geom_smooth(method = "lm", se = FALSE, color = "#3498DB") +
    labs(
      title = "Scatterplot of Education vs Literacy with Regression Line",
      x = "Literacy Rate (%)",
      y = "College enrollment (Y)"
    ) +
    theme_minimal()

```

# Regression

Use base R to run a regression.

```{r}
#| label: regression

lm(educ ~ literacy, data = df)

```

Run a correlation test.

```{r}
#| label: cor.test

cor.test(
     ~ educ + literacy,
     data = df,
     method = "pearson",
     na.action = na.omit
 ) 

```

Use `get_regression_table()` to run a regression.

```{r}
#| label: regress-table

model <- lm(educ ~ literacy, data = df)

get_regression_table(model)

```

# Categorical IV

Add a categorical variable to the example df.

```{r}
#| label: cat-var

# Use mutate and case_when to add the country_group column
df <- df %>%
  mutate(country_group = case_when(
    literacy < 50 ~ "Low Literacy",
    literacy >= 50 & literacy <= 80 ~ "Medium Literacy",
    literacy > 80 ~ "High Literacy"
  ))

head(df)

```

Use `get_regression_table()` with a categorical IV.

```{r}
#| label: lm-2

model_2 <- lm(educ ~ country_group, data = df)

get_regression_table(model_2)
```

# $R^2$

Use `get_regression_points()` to see the residuals.

```{r}
#| label: get_regression_points

model <- lm(educ ~ literacy, data = df)

get_regression_points(model)

```

Calculate $R^2$

```{r}
#| label: r-squared

my_test <- cor.test(
     ~ educ + literacy,
     data = df,
     method = "pearson",
     na.action = na.omit
 ) 

my_test$estimate
(my_test$estimate)^2

```

Use `get_regression_summaries()` to see $R^2$.

```{r}
#| label: get-regression-summ

model <- lm(educ ~ literacy, data = df)

get_regression_summaries(model)
```
