---
title: "Code-along 02"
author: "FirstName LastName"
editor: visual
execute:
  error: true
format: pdf # or replace 'pdf' with 'typst'
---


# Setup

## Install new package

This chunk will not run (on purpose!) when rendered because
`eval: false`.

**Install the
[`summarytools`](https://cran.r-project.org/web/packages/summarytools/vignettes/introduction.html)
package, available on CRAN.** **Click the green arrow icon to run the
following code chunk.**

```{r}
#| eval: false
#| message: false
#| label: install-summarytools

install.packages("summarytools")
```

## Load the packages

Load the standard packages and our new package `summarytools()`.

```{r}
#| message: false
#| label: load

library(tidyverse) 
library(haven) # not core tidyverse
library(gssr)
library(gssrdoc)
library(summarytools)
```

## Load your data

```{r}
#| message: false
#| label: gssdata

# Get the data only for the 2024 survey respondents
gss24 <- gss_get_yr(2024)

# Get the data all survey years
data(gss_all)

```

Use the `gss_which_years()` function to identify the survey years a variable appeared.

```{r}
#| label: which-years

gss_which_years(gss_all, meovrwrk)

```

**Which survey years did the variable `childs` appear?**

```{r, eval=FALSE}
#| label: which-years-mini-task

# REPLACE THE BLANK LINE WITH THE NAME OF THE VARIABLE

gss_which_years(gss_all, ________)

```


# Variable Types

## Data types in R

A property is assigned to objects that determines how generic functions
operate with it.

**logical** - Boolean values `TRUE` and `FALSE`

```{r}
class(TRUE)
```

**character** - character strings

```{r}
class("Sociology")
```

**Integer** - numeric data without decimals\
(indicated with an `L`).

```{r}
class(2L)
```

**numeric** - default type if values are numbers or if the values
contain decimals.

```{r}
class(2.5)
```

**factors** consist of character data with a fixed and known set of
possible values

```{r}
#| label: factors01

opinion <- factor(c("like", "dislike", "dislike", "hate", "dislike", "hate"))
class(opinion)

# By default, the levels are sorted alphabetically. 
levels(opinion) 

```

```{r}
#| label: factors02

# Reorder the levels with the argument `levels` in the `factor()` function
opinion <- factor(opinion, levels = c("hate", "dislike", "like")) 
levels(opinion)
```

```{r}
#| label: factors03

# If the order has meaning (like rankings), you can make it an ordered factor
opinion <- factor(opinion, levels = c("hate", "dislike", "like"), ordered = TRUE) 
levels(opinion)
```

## Converting between types

Use a function: `as.logical()`, `as.numeric()`, `as.integer()`, or
`as.character()`.

Create a numeric variable.

```{r}
#| label: convert_01

x <- 1:3
x
class(x)
```

Change it to a character variable.

```{r}
#| label: convert_02

y <- as.character(x)
y
class(y)
```

## Haven labelled

When you import data into R from software like SPSS, Stata, or SAS, you
might notice a special class called `haven_labelled`.

```{r}
#| label: labelled

class(gss24$premarsx)
table(gss24$premarsx)
```

It makes data easier to understand without needing a separate codebook.

```{r}
#| label: variable_labels

attr(gss24$premarsx, "label") 

```

See the value labels with `print_labels()`

```{r}
#| label: value_labels

print_labels(gss24$premarsx) 

```

You can use
[`as_factor`](https://haven.tidyverse.org/reference/as_factor.html) to
see the value labels of the variable `premarsx`.

```{r, eval=FALSE}
#| label: label02_premarsx

# REPLACE THE BLANK LINE WITH THE NAME OF THE VARIABLE

table(as_factor(gss24$________), useNA = "ifany")

```

## Convert labels to factors

1.  Get rid of all the 'missing' (NA) levels using `zap_missing`

```{r, eval=FALSE}
#| label: label03_premarsx

# REPLACE THE BLANK LINES WITH THE NAME OF THE VARIABLE

gss24$premarsx <- zap_missing(gss24$________) 

table(as_factor(gss24$________), useNA = "ifany") # see the results
```

2.  Apply the labels instead of numeric values using `as_factor`

```{r, eval=FALSE}
#| label: label04_premarsx

# REPLACE THE BLANK LINE WITH THE FUNCTION `as_factor()`

gss24$premarsx <- ________(gss24$premarsx) # replaces the values with labels
table(gss24$premarsx, useNA = "ifany") # no longer need to wrap the variable in as_factor
```

3.  Get rid of the empty levels in `premarsx` using `droplevels`

```{r, eval=FALSE}
#| label: label05_premarsx

# REPLACE THE BLANK LINES WITH THE `droplevels()` function & the NAME OF THE VARIABLE

gss24$________ <- __________(gss24$________)
table(gss24$________)
```

Now, use `zap_missing()`, `as_factor()`, and `droplevels()` to do the
same for the `sex` variable.

Then, use `table()` to see your results.

```{r}
#| label: labels_sex

# ADD YOUR CODE HERE
```

# Frequency Distributions

Make a frequency table of the variable `sex`.

```{r}
#| label: freq01

freq(gss24$sex) 

```

Make a pretty frequency table for the variable `premarsx`.

```{r}
#| label: freq02

# ADD YOUR CODE HERE

```

Using `report.nas = FALSE` suppresses the missing data.\
The `headings = FALSE` parameter suppresses the heading section. Do the
same for `premarsx`.

```{r}
#| label: freq03

freq(gss24$sex, report.nas = FALSE, headings = FALSE) 

# ADD YOUR CODE HERE
```

Use these options with the `freq()` function for the variable
`premarsx`.

```{r}
#| label: freq04

# ADD YOUR CODE HERE

```

## Cross-tabs

We've been using the `table()` function with one variable at a time, but
it also let's you create a frequency table (**crosstab**) with two
variables.

```{r}
#| label: crosstab01

# 1st variable is the rows, 2nd variable is the columns.
table(gss24$premarsx, gss24$sex)

```

To run `freq()` by group, pair it with the `stby()` function.

```{r}
#| label: crosstab02

stby(gss24$premarsx, gss24$sex, freq)
```

Use `summarytools::ctable` instead!

```{r}
#| label: crosstab03

ctable(gss24$premarsx, gss24$sex, 
       prop = "c", 
       format = "p", 
       useNA = "no")  
```

## Show your knowledge

Based on your table:

- ***\[your answer here\]*** percentage of respondents believe sex
  before marriage is 'almost always wrong'?

- A greater percentage of ***\[men or women\]*** think sex before
  marriage is 'not wrong at all'.

# Data Management I

## The pipe `|>`

The pipe operator passes what comes before it into the function that comes after it as the first argument in that function.

```{r}
sum(1, 2)
```

```{r}
1 |> 
  sum(2)
```

## `dplyr()` in action

Create a frequency table for `childs` variable.

```{r}
#| label: freq-tb

gss_all$childs <- zap_missing(gss_all$childs) 
gss_all$childs <- as_factor(gss_all$childs)   
gss_all$childs <- droplevels(gss_all$childs)  

gss_all |> 
  freq(childs, report.nas = FALSE) |> 
  tb() 

```

**Use a `|>` & `gss_which_years()` to find out which survey years the variable `agekdbrn` appears.**

```{r, eval=FALSE}
#| label: which-years-pipe

# ADD YOUR CODE HERE

```

## `filter()` & `drop_na()` {.sparse-slide}

Use `filter()` to keep rows that meet a condition.\
Use `drop_na()` to remove rows with missing (NA) values.

```{r}
#| label: pipe01

gss_all |>
  filter(year == 2022) |>
  drop_na(sex, agekdbrn) |>
  count(sex)

```

<<<<<<< HEAD
**Limit the dataframe to only the 2024 respondents without missing data for `sex`, `hrs1`, and `childs`. How many men and women are left?**
=======
**Limit the dataframe to only the 2024 respondents without missing data
for `sex`, `hrs1`, and `childs`. How many men and women are left?**
>>>>>>> origin/main

```{r, eval=FALSE}
#| label: pipe01-mini-task

gss_all |>
  filter(year == 2022)  # CHANGE THE YEAR TO 2024 & ADD A PIPE OPERATOR 
  drop_na() |> # ADD THE VARIABLES INSIDE drop_na()
  count(sex)

```

# Summary Statistics

## Median & Mean w base R

```{r}
#| label: median

# Using base R
median(gss_all$agekdbrn, na.rm=TRUE) 
```

Note: `na.rm` is a logical evaluating to `TRUE` or `FALSE` indicating whether `NA` values should be stripped before the computation proceeds.

```{r}
#| label: mean

mean(gss_all$agekdbrn, na.rm=TRUE)
```

```{r}
#| label: summary

# show me summary statistics
summary(gss_all$agekdbrn)
```

**Find the mean and median for the variable: `hrs1`.**

```{r, eval=FALSE}
#| label: summary-mini-task

# REPLACE THE BLANK LINES WITH THE DATA FRAME NAME & VARIABLE NAME
median(______________, na.rm=TRUE)
mean(______________, na.rm=TRUE)

# USE summary() TO SHOW SUMMARY STATISTICS


```

## `summarise()` {.sparse-slide}

Use functions inside `summarise()` to create summary variables.

```{r}
#| label: pipe02-a

gss_all |>  
  drop_na(agekdbrn) |>
   summarise(
    avg = mean(agekdbrn)
    ) 
    
```

## Median & Mean with `dplyr` {.sparse-slide}

Use [dplyr grammar]() to find the frequency, mean, and median.

Add lots of summary variables using functions inside `summarise()`.

1.  Use `n()` function to create a count variable (no variable in parentheses)
2.  Use `mean()` function on a variable
3.  Use `median()` function on a variable

```{r}
#| label: pipe02-b

gss_all |>  
  drop_na(agekdbrn) |> 
  summarise(
    freq = n(), 
    med = median(agekdbrn),
    avg = mean(agekdbrn)
    ) 

```

**Use `dplyr` grammar to find the frequency, mean, and median for the variable: `hrs1`.**

```{r, eval=FALSE}
#| label: pipe02-mini-task

gss_all |>  
  drop_na(hrs1) |>
  summarise(
   # ADD TO THE PIPE HERE WITH YOUR CODE HERE
)
```

## `group_by()` and `summarize()` {.sparse-slide}

A common pipeline is `group_by()` and then `summarize()` to calculate summary statistics for each group:

Use `group_by()` to organize your data into groups based on one or more variables.\
Use `summarize()` to compute statistics like total, mean, or median for each group.\
<br>

1.  Start with the `gss_all` data frame:
2.  Remove any observations with missing data for our key variables
3.  Do the next steps separately for each group in the variable
4.  Creates a new data frame with one row for each combination of grouping variables

```{r}
#| label: pipe03

gss_all |>  
  drop_na(sex, agekdbrn) |> 
  group_by(sex) |> 
  summarise(avg = mean(agekdbrn)) 

```

**Compare the average and median work hours (`hrs`) for U.S. men and women.**

```{r, eval=FALSE}
#| label: pipe03-mini-task
#| code-line-numbers: "1-3|4|5-7"
#| output-location: column-fragment

gss_all |>  
  drop_na(sex, hrs1) |> 
   # ADD TO THE PIPE HERE WITH YOUR CODE HERE

```

## `summarize()` + variability measures {.sparse-slide}

```{r}
#| label: summarize

gss_all |>  
  drop_na(agekdbrn, sex)  |>
  group_by(as_factor(sex)) |>  
  summarise(
    freq = n(),
    min = min(agekdbrn),
    median = median(agekdbrn),
    max = max(agekdbrn),
    mean = round(mean(agekdbrn), digits = 2), 
    sd = sd(agekdbrn)
    ) 

```

# Think Like a Statistician

**How did the mean and standard deviation of weekly work hours differ between men and women in 2024, and what do these differences suggest about gender-based variability in labor patterns.**

*How do we find out?*

```{r}
#| label: think-output

# ADD YOUR CODE HERE

```

## Your Data Take

**What's your conclusion to our initial research question?**
