---
title: "Code-along 03"
author: "FirstName LastName"
editor: visual
execute:
  error: true
format: pdf # or replace 'pdf' with 'typst'
---

# Setup

## Packages

Load the standard packages.

```{r}
#| message: false
#| label: load

library(tidyverse) 
library(haven) # not core tidyverse
library(gssr)
library(gssrdoc) # load the GSS codebook
library(summarytools)
```

## Load your data & codebook

```{r}
#| message: false
#| label: gssdata

# Get the data all survey years
data(gss_all)

```

Use the `gss_which_years()` function to identify the survey years a
variable appeared.

```{r}
#| label: which-years

gss_which_years(gss_all, meovrwrk)

```

**Which survey years did the variable `childs` appear?**

```{r, eval=FALSE}
#| label: which-years-mini-task

# REPLACE THE BLANK LINE WITH THE NAME OF THE VARIABLE

gss_which_years(gss_all, ________)

```

# Data Management I

## The pipe `|>`

The pipe operator passes what comes before it into the function that
comes after it as the first argument in that function.

```{r}
sum(1, 2)
```

```{r}
1 |> 
  sum(2)
```

## `dplyr()` in action

Create a frequency table for `childs` variable.

```{r}
#| label: freq-tb

gss_all$childs <- zap_missing(gss_all$childs) 
gss_all$childs <- as_factor(gss_all$childs)   
gss_all$childs <- droplevels(gss_all$childs)  

gss_all |> 
  freq(childs, report.nas = FALSE) |> 
  tb() 

```

**Use a `|>` & `gss_which_years()` to find out which survey years the
variable `agekdbrn` appears.**

```{r, eval=FALSE}
#| label: which-years-pipe

# ADD YOUR CODE HERE

```

## `filter()` & `drop_na()` {.sparse-slide}

Use `filter()` to keep rows that meet a condition.\
Use `drop_na()` to remove rows with missing (NA) values.

```{r}
#| label: pipe01

gss_all |>
  filter(year == 2022) |>
  drop_na(sex, agekdbrn) |>
  count(sex)

```

**Limit the dataframe to only the 2024 respondents without missing data for `sex`, `hrs1`, and `childs`. How many men and women are left?**

```{r, eval=FALSE}
#| label: pipe01-mini-task

gss_all |>
  filter(year == 2022)  # CHANGE THE YEAR TO 2024 & ADD A PIPE OPERATOR 
  drop_na() |> # ADD THE VARIABLES INSIDE drop_na()
  count(sex)

```

# Summary Statistics

## Median & Mean w base R

```{r}
#| label: median

# Using base R
median(gss_all$agekdbrn, na.rm=TRUE) 
```

Note: `na.rm` is a logical evaluating to `TRUE` or `FALSE` indicating
whether `NA` values should be stripped before the computation proceeds.

```{r}
#| label: mean

mean(gss_all$agekdbrn, na.rm=TRUE)
```

```{r}
#| label: summary

# show me summary statistics
summary(gss_all$agekdbrn)
```

**Find the mean and median for the variable: `hrs1`.**

```{r, eval=FALSE}
#| label: summary-mini-task

# REPLACE THE BLANK LINES WITH THE DATA FRAME NAME & VARIABLE NAME
median(______________, na.rm=TRUE)
mean(______________, na.rm=TRUE)

# USE summary() TO SHOW SUMMARY STATISTICS


```

## `summarise()` {.sparse-slide}

Use functions inside `summarise()` to create summary variables.

```{r}
#| label: pipe02-a

gss_all |>  
  drop_na(agekdbrn) |>
   summarise(
    avg = mean(agekdbrn)
    ) 
    
```

## Median & Mean with `dplyr` {.sparse-slide}

Use [dplyr grammar]() to find the frequency, mean, and median.

Add lots of summary variables using functions inside `summarise()`.

1.  Use `n()` function to create a count variable (no variable in
    parentheses)
2.  Use `mean()` function on a variable
3.  Use `median()` function on a variable

```{r}
#| label: pipe02-b

gss_all |>  
  drop_na(agekdbrn) |> 
  summarise(
    freq = n(), 
    med = median(agekdbrn),
    avg = mean(agekdbrn)
    ) 

```

**Use `dplyr` grammar to find the frequency, mean, and median for the
variable: `hrs1`.**

```{r, eval=FALSE}
#| label: pipe02-mini-task

gss_all |>  
  drop_na(hrs1) |>
  summarise(
   # ADD TO THE PIPE HERE WITH YOUR CODE HERE
)
```

## `group_by()` and `summarize()` {.sparse-slide}

A common pipeline is `group_by()` and then `summarize()` to calculate
summary statistics for each group:

Use `group_by()` to organize your data into groups based on one or more
variables.\
Use `summarize()` to compute statistics like total, mean, or median for
each group.\
<br>

1.  Start with the `gss_all` data frame:
2.  Remove any observations with missing data for our key variables
3.  Do the next steps separately for each group in the variable
4.  Creates a new data frame with one row for each combination of
    grouping variables

```{r}
#| label: pipe03

gss_all |>  
  drop_na(sex, agekdbrn) |> 
  group_by(sex) |> 
  summarise(avg = mean(agekdbrn)) 

```

**Compare the average and median work hours (`hrs`) for U.S. men and
women.**

```{r, eval=FALSE}
#| label: pipe03-mini-task
#| code-line-numbers: "1-3|4|5-7"
#| output-location: column-fragment

gss_all |>  
  drop_na(sex, hrs1) |> 
   # ADD TO THE PIPE HERE WITH YOUR CODE HERE

```

## `summarize()` + variability measures {.sparse-slide}

```{r}
#| label: summarize

gss_all |>  
  drop_na(agekdbrn, sex)  |>
  group_by(as_factor(sex)) |>  
  summarise(
    freq = n(),
    min = min(agekdbrn),
    median = median(agekdbrn),
    max = max(agekdbrn),
    mean = round(mean(agekdbrn), digits = 2), 
    sd = sd(agekdbrn)
    ) 

```

# Think Like a Statistician

**How did the mean and standard deviation of weekly work hours differ
between men and women in 2024, and what do these differences suggest
about gender-based variability in labor patterns.**

*How do we find out?*

```{r}
#| label: think-output

# ADD YOUR CODE HERE

```

## Your Data Take

**What's your conclusion to our initial research question?**
