---
title: "Milestone 03"
author: "FirstName LastName"
format: pdf # or replace 'pdf' with 'typst'
execute:
  error: true # change this to false when your assignment is complete.
---

# Environment

Update the YAML with your first and last name.

Load the packages.

```{r}
#| label: load
#| message: false

library(tidyverse) 
library(haven) 
library(gssr)
library(gssrdoc)
library(summarytools)
library(Rmisc) # load the new package
library(gtsummary) # load the new package
```

Load the 2022 GSS data

```{r}
#| label: gss-data
#| message: false

# ADD YOUR CODE HERE TO LOAD THE GSS 2022 DATA 
gss22 <- gss_get_yr(2022)

```

You’ll be working with the following GSS variables:

-   `id`: Respondent id
-   `meovrwrk`: Men hurt family when focus on work too much
-   `sex`: Respondent’s gender
-   `lifenow` R's rating of life overall now from 0-10
-   `educ`  Highest year of school completed

# Code, Output, Meaning

*Use R to complete the checkpoints below. Show your work (e.g., R code
chunks) where appropriate. Add narrative (text outside code chunks) or
comments (text inside the code chunks) throughout.*

Reference specific statistics (where appropriate) from your output to
justify your answers. Explain what the values tell you about the data;
interpret their meaning in relation to the question.

## Checkpoint 01: Create a new dataframe

A.  Create a df named `my_data` with `meovrwrk`, `sex`, `marital`, `lifenow`, and `educ`. Then, use `drop_na()` to remove rows with missings data.

```{r}
#| label: my_data

my_data <- gss22 |>
  select(id, meovrwrk, sex, marital, lifenow, educ) |>
  drop_na()

```


## Checkpoint 02: Clean and create variables

A.   Use `mutate()` and `case_when()` to add a new variable to `my_data`
    called `hurt_num` that groups respondents as follows:
    
-   1 if `meovrwrk` is 'strongly agree' or 'agree'
-   0 if `meovrwrk` equal anything else

```{r}
#| label: hurtovrwrk

my_data <- my_data |>
  mutate(
    hurt_num = case_when(
      meovrwrk %in% 1:2 ~ 1,
      TRUE ~ 0 
    )) 

table(my_data$meovrwrk, my_data$hurt_num)

```

B.  Use `mutate()` and `case_when()` to add a new variable to `my_data`
    called `college` that groups respondents as follows:
    
-   "No college degree" if `educ` is less than 16
-   "College degree" if `educ` 16 or more

```{r}
#| label: college

my_data <- my_data |>
  mutate(
    college = case_when(
      educ < 16 ~ "No college degree",
      educ >= 16 ~ "College degree"
  )) 

table(my_data$educ, my_data$college)

```


C.  Use `mutate()` and `case_when()` to add a new variable to `my_data`
    called  `edu_category` that groups respondents as follows:
    
-   "High school or less" if `educ` is 12 or less 
-   "Some college" if `educ` is equal to or greater than 13 and equal to or less than 15
-   "College or more" if `educ` is 16 or more

```{r}
#| label: educ

my_data <- my_data |>
  mutate(
    edu_category = case_when(
      educ <= 12 ~ "High school or less",
      educ >= 13 & educ <= 15 ~ "Some college",
      educ >= 16  ~ "College or more"
  )) 

table(my_data$educ, my_data$edu_category)

```


D. Use `mutate()` and `case_when()` to add a new variable to `my_data`
    called  `marstat` that groups respondents as follows:
    
-   "Married" if `marital` is 'married'
-   "Single" if `marital` is 'never married'
-   "Prev married" if `marital` is 'widowed', 'divorced', or 'separated'

```{r}
#| label: educ

my_data <- my_data |>
  mutate(
    marstat = case_when(
      marital == 1 ~ "Married",
      marital == 5 ~ "Single",
      marital %in% 2:4   ~ "Prev married"
  )) 

table(my_data$marital, my_data$marstat)

```

## Checkpoint 03: Create and interpret confidence intervals

A. Create 95% confidence interval for `lifenow`

```{r}
#| label: ci-lifenow-95

my_data |>
  summarise(
    n = n(),
    mean_lifenow = mean(lifenow),
    sd_lifenow = sd(lifenow),
    lowCI_lifenow = CI(lifenow, ci = 0.95)['lower'],
    hiCI_lifenow  = CI(lifenow, ci = 0.95)['upper']
    )

```

**Interpret your confidence interval (e.g.,say in plain language what the CI is telling us).**

B. Create 99% confidence interval for `lifenow`

```{r}
#| label: ci-lifenow-99

my_data |>
  summarise(
    n = n(),
    mean_lifenow = mean(lifenow),
    sd_lifenow = sd(lifenow),
    lowCI_lifenow = CI(lifenow, ci = 0.99)['lower'],
    hiCI_lifenow  = CI(lifenow, ci = 0.99)['upper']
    )

```

*How does the 95% confidence interval differ from the 99% confidence interval in terms of width and certainty?*

## Checkpoint 04: Create, compare and interpret confidence intervals

A. Create 95% confidence intervals for `lifenow` by `college`

```{r}
#| label: ci-lifenow-college

my_data |>
  group_by(college) |>
  summarise(
    n = n(),
    mean = mean(lifenow),
    sd = sd(lifenow),
    lowCI = CI(lifenow, ci = 0.95)['lower'],
    hiCI  = CI(lifenow, ci = 0.95)['upper']
    )

```

**Do the confidence intervals for these two groups overlap? What does that imply about the statistical significance of their difference?**

B. Create 95% confidence intervals for `lifenow` by `edu_category`

```{r}
#| label: ci-lifenow-college

my_data |>
  group_by(edu_category) |>
  summarise(
    n = n(),
    mean = mean(lifenow),
    sd = sd(lifenow),
    lowCI = CI(lifenow, ci = 0.95)['lower'],
    hiCI  = CI(lifenow, ci = 0.95)['upper']
    )

```

**Do the confidence intervals for these three groups overlap? What does that imply about the statistical significance of their difference?**

## Checkpoint 05: Create, compare and interpret confidence intervals

A. Create 95% confidence intervals for `hurt_num` by `sex` & `marstat` (e.g, married women, married men, single women, single, men, etc.)

```{r}
#| label: ci-hurt_num-marstat

my_data |>
  group_by(sex, marstat) |>
  summarise(
    n = n(),
    mean = mean(hurt_num),
    sd = sd(hurt_num),
    lowCI = CI(hurt_num, ci = 0.95)['lower'],
    hiCI  = CI(hurt_num, ci = 0.95)['upper']
    )

```


## Checkpoint 06: Write a null and research hypothesis


## Checkpoint 07: Produce and interpret a two-sample t-tests

lifenow & college


## Checkpoint 08: Create cross-tabs with chi-square tests

Use `tbl_cross()` and `add_p()` to create a cross-tab with a chi-square test to answer the following questions.


A. Is there a statistically significant gender difference (`sex`) in agreement that men hurt the family when they focus on work too much (`meovrwk`)?

```{r}
#| message: false
#| label: tbl-cross-meovrwrk

my_data |>
  tbl_cross(
    row = meovrwrk,
    col = edu_category,
    percent = "column") |>
  add_p(source_note = TRUE)
```


## Checkpoint 09: Produce and interpret a Pearson’s Correlation Coefficient (R)

lifenow & educ


## Checkpoint 10:

# IPUMS Data

To keep your Research Brief progress on track, you'll complete short
exercises that correspond with the new course material using your own
dataset as part of your milestones.

-   Choose an interval-ratio (e.g, continuous or proportion) variable that you wish to consider as a **dependent variable** (DV).

-   Create a summary table for this DV, including the total number of observations, mean, median, and mode.
    
-    Pick or create a dichotomous variable (e.g., gender, two racial categorizes; two age categories; two countries; two years; etc.) in your dataset and create confidence intervals for each group for your DV. Compare and interpret the confidence intervals.

-   Write a null and research hypothesis using these two variables.

-   Conduct a two sample t-test (using `t.test()`) using your DV and dichotomous variable. Interpret your results. How do they relate to your hypotheses?

-   Choose a second interval-ratio variable that you think may be associated with your DV. 

-   Use `cor.test()` to formally test for an association between the two continuous variables. Interpret your results.
