---
title: "Milestone 01"
author: "FirstName LastName"
format: pdf
---

# Environment

Update the YAML with your first and last name.

Load the packages from the code-along. (Install first if not already
done so.)

```{r}
#| label: load
#| message: false

library(here)
library(tidyverse) 
library(haven) 
library(gssr)
library(gssrdoc)
library(summarytools)
```

# Add Documentation

Run the code and add **narrative** (outside code chunks) or **comments**
(inside the code chunks) describing what each of the the below R chunks
do.

```{r}
#| label: here

here()
```

```{r}
#| label: list-files

list.files(path = here())

list.files(path = here("project"))
```

```{r}
#| label: time

Sys.time()
```

# Explore the GSS

Load all years of the gss data.

Then, use the `gss_get_yr()` function to load the data only for the 2024
survey respondents. (*Hint*: Use the code from code-along-01.qmd)

```{r}
#| label: gss-datasets

# Load the data (will appear in your Global Environment pane)
data(gss_all)
 
# load the gss 2024 data (add your code below)

```

Use R to answer the questions below. Show your work (e.g., R code
chunks) where appropriate. Add narration and comments throughout.
Justify your answers.

-   Report the survey question that corresponds to the `fefam` variable.

-   Is `fefam` a dichotomous, nominal, ordinal, or interval-ratio
    variable? Justify your answer.

-   Which years of the survey did the `fefam` variable appear?

-   What value corresponds with "agree" for this variable?

-   Convert the fefam variable to a factor variable. Make sure you
    remove any 'missing' (NA) or empty (e.g., levels with no
    observations) levels.

-   Create a (pretty) relative frequency table to show for the `fefam`
    variable using the 2024 data. How many respondents "strongly agree"
    with the `fefam` statement in 2024? What % of respondents either
    "strongly agree" or "agree"?

-   Create a (pretty) relative frequency table for the `happy` variable
    using the 2024 data. Make sure your values are labelled for easy
    interpretation (without levels identifying missing data or empty
    levels). Were most respondents "not too happy," "pretty happy," or
    "very happy" in 2024?

-   Create a (pretty) cross-tab of the `happy` and `sex` variables using
    only the 2024 survey data. Make sure your values are labelled for
    easy interpretation. How many women were `very happy`? How many men
    were `very happy`?

-   Were a greater percentage of women or men `very happy` in 2024?

# Meet the IPUMS Data

To keep your Research Brief progress on track, you'll complete short
exercises that correspond with the new course material using your own
dataset as part of your milestones.

Consider a research topic you are interested in and explore the
documentation for some of the available [IPUMS (microdata)
datasets](https://www.ipums.org/overview) that could be relevant for the
subject. Start broadly, such as looking at the [Topical Availability
Summary](https://www.ipums.org/topics). Then, look at a couple of
individual datasets more in depth. Explore the sidebar of available
documentation for 2-3 datasets: About, Data, Documentation (including
User Guides and Questionnaires, FAQs).

-   Write 1-2 sentences about your research interests.

-   Then, write 2-3 sentences about which IPUMS dataset you are thinking
    of using, explaining the sample population and time frame of the
    dataset, and why it is a good fit for your research interests.

-   Describe 2-3 key variables available in the dataset that are a good
    fit for your research topic. Report the exact survey questions and
    response options for these variables.

-   Finally, create an IPUMS account for the dataset of interest. See
    this [help video](https://www.youtube.com/watch?v=Oo7RYMV7q4Y) for
    details.

**NOTE:** For your Research Brief, you must choose an [IPUMS (microdata)
dataset](https://www.ipums.org/overview) to use.
