---
title: "Milestone 01"
author: "FirstName LastName"
format: html
---

# Environment

Update the YAML with your first and last name.

Load the packages from the code-along. (Install first if not already done so.)

```{r}
#| label: load

library(here)
library(tidyverse)
library(gssr)
library(gssrdoc)
```

# Add Documentation

Run the code and add **narrative** (outside code chunks) or **comments** (inside the code chunks) to the R chunks below.

```{r}
#| label: here

here()
```

```{r}
#| label: list-files

list.files(path = here())

list.files(path = here("project"))
```

```{r}
#| label: time

Sys.time()
```

# Explore the GSS

Load all years of the gss data. Then, use the `gss_get_yr()` command to load the data only for the 2024 survey respondents.

*Hint*: Use the code from code-along-01.qmd

```{r}
#| label: gss-datasets

# Load the data (will appear in your Global Environment pane)
data(gss_all)
 
# load the gss 2024 data (add your code below)

```

Use R to answer the questions below. Show your work (e.g., R code) when appropriate. Add narration and comments throughout.

-   Report the survey question that corresponds to the `fefam` variable.

-   What type is the `fefam` variable? (nominal, ordinal, interval-ratio, dichotomous)? Justify your answer.

-   Which years of the survey did the `fefam` variable appear?

-   What value corresponds with "agree" for this variable?

-   How many people "strongly agree" with the `fefam` statement in 2024?

-   Create a crosstab of the `happy` and `sex` variables using only the 2024 survey data. How many women were `very happy`? How many men were `very happy`?

# Meet the IPUMS Data

<https://www.ipums.org/overview>

<https://cran.r-project.org/web/packages/ipumsr/vignettes/ipums.html>

<https://tech.popdata.org/ipumsr/>
