---
title: "Milestone 99"
author: "FirstName LastName"
format: pdf # or replace 'pdf' with 'typst'
execute:
  error: true # change this to false when your assignment is complete.
---

This assignment asks students to work with wide and long dataformats.

# Environment

Update the YAML with your first and last name.

Load the packages.

```{r}
#| label: load
#| message: false

library(tidyverse) 
library(haven) 
library(gssr)
library(gssrdoc)
library(summarytools)
library(Rmisc) # load the new package
library(gtsummary) # load the new package
```

Load the GSS panel data

```{r}
#| label: gss-data
#| message: false

# Use here() to construct the file path of the panel data
gss_panel.dta <- here("data", "GSS_2020_panel_stata_1a/gss2020panel_r1a.dta")

#load the panel data using `haven::read_dta()`
gss_panel <- read_dta(gss_panel.dta)

```


*Use the panel data for checkpoints 06-10.*

## Checkpoint 06: Code for within-person change

Run the following code to produce the average within-person change
between 2018 and 2020.

It shows the average change from 2018 to 2020 in work and TV hours for
each gender.

Positive values mean an increase from 2018 to 2020; negative values mean
a decrease.

```{r}
#| label: within-change

gss_panel |>
  select(sex_2, hrs1_2, hrs1_1b, tvhours_2, tvhours_1b) |>
  drop_na(sex_2) |>
  mutate(
    work_change = hrs1_2 - hrs1_1b,
    tv_change = tvhours_2 - tvhours_1b,
  ) |>
  group_by(as_factor(sex_2)) |>
  summarise(
    avg_work_change = round(mean(work_change, na.rm = TRUE), digits = 2),
    sd_work_change = round(sd(work_change, na.rm = TRUE), digits = 2),
    avg_tv_change = round(mean(tv_change, na.rm = TRUE), digits = 2),
    sd_tv_change = round(sd(tv_change, na.rm = TRUE), digits = 2)
  )
```

***Did the number of work hours increase or decrease from 2018 to 2020
for men? What about for women?***

***Did the number of hours spent watching TV increase or decrease from
2018 to 2020 for men? What about for women?***

***Was there a gender difference in the average within-person change
between 2018 and 2020 for work hours? What about for tv hours?***

## Checkpoint 07: Reshape panel data so it is tidy

Use `select()` and `pivot_longer()` to create tidy data.

Add `relocate()` to put the variables in a logical order and `drop_na()`
to remove rows with any missing data.

```{r}
#| label: tidy-data

my_gss_panel <- gss_panel |>
  select(yearid, sex_2,
         starts_with("hrs1_"),
         starts_with("tvhours_")) |>
    pivot_longer(
        cols = c( -yearid, -sex_2),
        names_to = c(".value", "panel"),
        names_sep = "_") |>
  relocate(panel, .after = yearid) |>
  drop_na()

head(my_gss_panel)
```

## Checkpoint 08: Recode panel variables

A.  Use `mutate()` and `case_when()` to re-code the `sex_2` variable so
    that "1" = man, "2" = woman. Use `TRUE ~ NA_character_` for missing
    values.

```{r}
#| label: recode-sex

my_gss_panel <- my_gss_panel |>
  mutate(sex = case_when(
         sex_2 == 1 ~ "male",
         sex_2 == 2 ~ "female",
         TRUE ~ NA_character_))

```

B.  Use `mutate()` and `case_when()` to re-code the `panel` variable so
    that "1a" = 2016, "1b" = 2018, and "2" = 2020. Use
    `TRUE ~ NA_integer_` for missing values.

```{r}
#| label: recode-panel

my_gss_panel <- my_gss_panel |>
  mutate(panel = case_when(
         panel == "1a" ~ 2016,
         panel == "1b" ~ 2018,
         panel == "2" ~ 2020,
         TRUE ~ NA_integer_))

head(my_gss_panel)
```

## Checkpoint 09: Create a summary panel table

Use `group_by()` and `summarise()` to create a table showing the average
work hours for men and women for each panel year.

```{r}
#| label: summary-panel

panel_summary <- my_gss_panel |>
  group_by(sex, panel) |>
    summarise(
    count = n(),
    avg_work = round(mean(hrs1), digits = 2), 
    avg_tv = round(mean(tvhours), digits = 2)
    ) 

panel_summary
```

***Do women differ in their average work and television hours from 2016,
2018, and 2020? What about men?***

## Checkpoint 10: Join and calculate the average panel differences by sex

Use `full_join()` to merge the `gss24_summary` data and the
`panel_summary` data, matching *by* the variable `sex`.

Calculate the difference in average work and tv hours between 2016,
2018, 2020 and the 2024 average.

```{r}
#| label: change

## ADD YOUR CODE HERE TO JOIN THE DATA
my_data_all <- full_join(gss24_summary, panel_summary, by = "sex")

# Create change variables
my_data_all <- my_data_all |>
  mutate(
    work_change = avg_work_24 - avg_work,
    tv_change = avg_tv_24 - avg_tv
    ) |>
  select(sex, panel, count, 
         avg_work_24, avg_work, work_change,
         avg_tv_24, avg_tv, tv_change
         )

my_data_all
```

***Speculate about whether there is evidence of a post-pandemic (2024)
shift in work or TV time. Are the patterns the same or different for men
and women?***