---
title: "Milestone 03"
author: "Answer Key"
format: pdf # or replace 'pdf' with 'typst'
execute:
  error: true # change this to false when your assignment is complete.
---

# Environment

Update the YAML with your first and last name.

Load the packages.

```{r}
#| label: load
#| message: false

library(tidyverse) 
library(haven) 
library(gssr)
library(gssrdoc)
library(summarytools)
library(Rmisc) # load the new package
library(gtsummary) # load the new package
```

Load the 2022 GSS data

```{r}
#| label: gss-data

# ADD YOUR CODE HERE TO LOAD THE GSS 2022 DATA 
gss22 <- gss_get_yr(2022)

```

You’ll be working with the following GSS variables:

-   `meovrwrk`: Men hurt family when focus on work too much
-   `sex`: Respondent’s gender
-   `lifenow` R's rating of life overall now from 0-10
-   `educ` Highest year of school completed

# Code, Output, Meaning

*Use R to complete the checkpoints below. Show your work (e.g., R code
chunks) where appropriate. Add narrative (text outside code chunks) or
comments (text inside the code chunks) throughout.*

***Show your df and variable changes as you go (e.g., `table()`,
`head()`).*** Reference specific statistics (where appropriate) from
your output to justify your answers. Explain what the values tell you
about the data; interpret their meaning in relation to the question.

## Checkpoint 01: Create a new dataframe

A.  Create a df named `my_data` with `meovrwrk`, `sex`, `marital`,
    `lifenow`, and `educ`. Then, use `drop_na()` to remove rows with
    missing data.

```{r}
#| label: my_data

my_data <- gss22 |>
  select(meovrwrk, sex, marital, lifenow, educ) |>
  drop_na()

head(my_data)

```

## Checkpoint 02: Clean and create variables

A.  Use `mutate()` and `case_when()` to add a new variable to `my_data`
    called `hurt_num` that groups respondents as follows:

-   1 if `meovrwrk` is 'strongly agree' or 'agree'
-   0 if `meovrwrk` is equal to anything else

```{r}
#| label: hurtovrwrk

my_data <- my_data |>
  mutate(
    hurt_num = case_when(
      meovrwrk %in% 1:2 ~ 1,
      TRUE ~ 0 
    )) 

table(my_data$meovrwrk, my_data$hurt_num) # check the recode
table(my_data$hurt_num) # view counts

```

B.  Use `mutate()` and `case_when()` to add a new variable to `my_data`
    called `college` that groups respondents as follows:

-   "No college degree" if `educ` is less than 16
-   "College degree" if `educ` 16 or more

```{r}
#| label: college

my_data <- my_data |>
  mutate(
    college = case_when(
      educ < 16 ~ "No college degree",
      educ >= 16 ~ "College degree"
  )) 

table(my_data$educ, my_data$college)
table(my_data$college)
```

C.  Use `mutate()` and `case_when()` to add a new variable to `my_data`
    called `edu_category` that groups respondents as follows:

-   "High school or less" if `educ` is 12 or less
-   "Some college" if `educ` is equal to or greater than 13 and equal to
    or less than 15
-   "College or more" if `educ` is 16 or more

**BONUS: Then, use `factor()` to put the categories into a meaningful
order**

```{r}
#| label: edu_category

my_data <- my_data |>
  mutate(
    edu_category = case_when(
      educ <= 12 ~ "High school or less",
      educ >= 13 & educ <= 15 ~ "Some college",
      educ >= 16  ~ "College or more"
  )) 

# Convert edu_category to an ordered factor
my_data <- my_data |>
  mutate(
    edu_category = factor(
      edu_category,
      levels = c("High school or less", "Some college", "College or more")
      ))

table(my_data$educ, my_data$edu_category)
table(my_data$edu_category)

```

D. Use `mutate()` and `case_when()` to add a new variable to `my_data`
called `marstat` that groups respondents as follows:

-   "Married" if `marital` is 'married'
-   "Single" if `marital` is 'never married'
-   "Prev married" if `marital` is 'widowed', 'divorced', or 'separated'

**BONUS: Then, use `factor()` to put the categories into a meaningful
order**

```{r}
#| label: marstat

my_data <- my_data |>
  mutate(
    marstat = case_when(
      marital == 1 ~ "Married",
      marital == 5 ~ "Single",
      marital %in% 2:4   ~ "Prev married"
  )) 

# Convert marstat to an ordered factor
my_data <- my_data |>
  mutate(
    marstat = factor(
      marstat,
      levels = c("Single", "Married", "Prev married")
      ))

table(my_data$marital, my_data$marstat)
table(my_data$marstat)
```

E. Use `zap_missing()`, `as_factor()`, and `droplevels()` to make `sex`
and `meovrwkr` into clean factor variables.

```{r}
#| label: sex-meovrwkr

my_data$sex <- zap_missing(my_data$sex)
my_data$sex <- as_factor(my_data$sex)
my_data$sex <- droplevels(my_data$sex)

my_data$meovrwrk <- zap_missing(my_data$meovrwrk)
my_data$meovrwrk <- as_factor(my_data$meovrwrk)
my_data$meovrwrk <- droplevels(my_data$meovrwrk)

```

## Checkpoint 03: Create and interpret confidence interval

A.  Calculate the mean and 95% confidence interval for `lifenow`

```{r}
#| label: ci-lifenow-95

my_data |>
  dplyr::summarise(
    n = n(),
    mean_lifenow = mean(lifenow),
    sd_lifenow = sd(lifenow),
    lowCI_lifenow = CI(lifenow, ci = 0.95)['lower'],
    hiCI_lifenow  = CI(lifenow, ci = 0.95)['upper']
    )

```

***Interpret your confidence interval (e.g., say in plain language what
the CI is telling us).***

The average life satisfaction score of 7.73 on a 0–10 scale, indicates
generally high levels of satisfaction.

There is a 95% probability that the true population mean falls between
7.64 and 7.81. Or: We are 95% confident the true average life
satisfaction score is between 7.64 and 7.81. OR: Based on the sample,
there is 95% certainty that the population average life sanctification
is between 7.64 and 7.81.

B.  Calculate the mean and 99% confidence interval for `lifenow`

```{r}
#| label: ci-lifenow-99

my_data |>
  dplyr::summarise(
    n = n(),
    mean_lifenow = mean(lifenow),
    sd_lifenow = sd(lifenow),
    lowCI_lifenow = CI(lifenow, ci = 0.99)['lower'],
    hiCI_lifenow  = CI(lifenow, ci = 0.99)['upper']
    )

```

***How does the 95% confidence interval differ from the 99% confidence
interval in terms of width and certainty?***

A 99% confidence interval of 7.61 to 7.83 life satisfaction points
offers greater certainty that the true population mean is captured, but
it is less precise than the narrower 95% interval of 7.64 to 7.81. The
99% interval is 0.05 life satisfaction points wider, reflecting that
increased confidence comes at the cost of precision, as the estimate
becomes less specific.

## Checkpoint 04: Create, compare and interpret confidence intervals

A.  Calculate the mean and 95% confidence intervals for `lifenow` by
    `college`

```{r}
#| label: ci-lifenow-college

my_data |>
  group_by(college) |>
  dplyr::summarise(
    n = n(),
    mean = mean(lifenow),
    sd = sd(lifenow),
    lowCI = CI(lifenow, ci = 0.95)['lower'],
    hiCI  = CI(lifenow, ci = 0.95)['upper']
    )

```

***Do the confidence intervals for these two groups overlap? What does
that imply about the statistical significance of their difference?***

The non-overlapping 95% confidence intervals indicate a likely
statistically significant difference in average life satisfaction, with
college graduates reporting notably higher average life satisfaction
(7.92) than those without a degree (7.56).

B. Calculate the mean and 95% confidence intervals for `lifenow` by
`edu_category`

```{r}
#| label: ci-lifenow-edu_category

my_data |>
  group_by(edu_category) |>
  dplyr::summarise(
    n = n(),
    mean = mean(lifenow),
    sd = sd(lifenow),
    lowCI = CI(lifenow, ci = 0.95)['lower'],
    hiCI  = CI(lifenow, ci = 0.95)['upper']
    )

```

***Do the confidence intervals for these three groups overlap? What does
that imply about the statistical significance of their difference?***

The 95% confidence intervals for high school or less (7.40–7.71) and
some college (7.39–7.75) overlap substantially, suggesting no
statistically significant difference in average life satisfaction
between those two groups. However, the interval for college or more
group (7.80–8.04) does not overlap with either of the other groups,
indicating a statistically significant and meaningfully higher level of
life satisfaction among those with a college degree or more.

## Checkpoint 05: Create, compare and interpret confidence intervals

Calculate the mean and 95% confidence intervals for `hurt_num` by
`marstat` & `sex` (e.g, married women, married men, single women, single
men, etc.)

```{r}
#| label: ci-hurt_num-marstat

my_data |>
  group_by(marstat, as_factor(sex)) |>
  dplyr::summarise(
    n = n(),
    mean = mean(hurt_num),
    sd = sd(hurt_num),
    lowCI = CI(hurt_num, ci = 0.95)['lower'],
    hiCI  = CI(hurt_num, ci = 0.95)['upper']
    )

```

***Assess whether gender differences within each marital status group
are likely to be statistically meaningful.***

Among singles and previously married individuals, the 95% confidence
intervals for men and women overlap, indicating no statistically
significant gender difference in beliefs about whether men hurt their
families by overworking. In contrast, the married group shows
non-overlapping intervals, suggesting a statistically significant
difference between married men and married women on this issue.

## Checkpoint 06: Write a null and research hypothesis

***Write a null and research hypothesis for the association between the
variables `lifenow` and `college`***

**H0:** There is no difference in average life satisfaction scores
between people with a college education and those without.

**H1:** People with a college education have higher average life
satisfaction scores than those without.

## Checkpoint 07: Produce and interpret a two-sample t-test

Use `t.test()` to produce a two-sided t.test of the variables `lifenow`
and `college`

```{r}
#| label: ttest

t.test(my_data$lifenow ~ my_data$college, alternative = "two.sided")

```

***Explain what the test reveals about the statistical differences in
average life satisfaction between the two groups. What do the results
suggest about the null and research hypotheses?***

The t-test result (t = 4.2025, p \< .001) indicates a statistically
significant difference in average life satisfaction between individuals
with and without a college education. Therefore, we reject the null
hypothesis.

## Checkpoint 08: Create cross-tabs with chi-square tests

Use `tbl_cross()` and `add_p()` to create a cross-tab with a chi-square
test for the variables `meovrwrk` and `college`

```{r}
#| label: meovrwrk-college

my_data |>
  tbl_cross(
    row = meovrwrk,
    col = college,
    percent = "column") |>
  add_p(source_note = TRUE)

```

***Is there a statistically significant difference in agreement that men
hurt the family when they focus on work too much by college degree?***

A chi-square test yielded a p-value of 0.3, indicating that the
relationship between gender and agreement with whether men hurt the
family when they overwork is not statistically significant. Therefore,
we fail to reject the null hypothesis that college degree and agreement
are independent.

## Checkpoint 09: Calculate and interpret a chi-square test

Use `tbl_cross()` and `add_p()` to create a cross-tab with a chi-square
test for the variables `meovrwrk` and `sex`

```{r}
#| label: meovrwrk-sex

my_data |>
  tbl_cross(
    row = meovrwrk,
    col = sex,
    percent = "column") |>
  add_p(source_note = TRUE)
```

***Is there a statistically significant gender difference in agreement
that men hurt the family when they focus on work too much?***

A chi-square test yielded a p-value less than .001, so we reject the
null hypothesis. There appears to be a statistically significant
relationship between gender and agreement with the statement that men
hurt the family when they overwork.

## Checkpoint 10: Produce and interpret a Pearson’s Correlation Coefficient (R)

Use `cor.test()` to test the correlation between `lifenow` & `educ`

```{r}
#| label: cor-educ01

cor.test(
  ~ lifenow + educ,
  data = my_data,
  method = "pearson",
  na.action = na.omit
)

```

***Describe the strength and direction of any correlation between
education and life state.***

The Pearson correlation test reveals a statistically significant ( p \<
.001) but weak positive relationship (r = 0.12) between education and
life satisfaction, suggesting that higher education levels are modestly
associated with greater life satisfaction.

# IPUMS Data

To keep your Research Brief progress on track, you'll complete short
exercises that correspond with the new course material using your own
dataset as part of your milestones.

-   Choose an interval-ratio (e.g, continuous or proportion) variable
    that you wish to consider as a **dependent variable** (DV).

-   Create a summary table for this DV, including the total number of
    observations, mean, median, and sd.

-   Pick or create a dichotomous variable (e.g., gender, two racial
    categories; two age categories; two countries; two years; etc.) in
    your dataset and create confidence intervals for each group for
    your DV. Compare and interpret the confidence intervals.

-   Write a null and research hypothesis using these two variables.

-   Conduct a two sample t-test (using `t.test()`) for your DV and
    dichotomous variable. Interpret your results. How do they relate to
    your hypotheses?

-   Choose a second interval-ratio variable that you think may be
    associated with your DV.

-   Use `cor.test()` to formally test for an association between the two
    continuous variables. Interpret your results.
