---
title: "milestone-00"
author: "FirstName LastName"
format: html
editor: visual
---

# Getting Started in RStudio & R

*Instructions to the instructor:*

- give general overview of RStudio

- introduction to Quarto (have students open mileston-00.qmd)

## Checkpoint 01

In the `Console`, type:

```         
print("Hello, World!)
```

Now, click the green arrow below to "`Run Current Chunk`".

```{r}
print("Hello, World")
```

## Checkpoint 02

Store the values in an object called `my_data`. Click the green arrow
below to "`Run Current Chunk`".

```{r}

my_data <- c(48000, 52000, 45000, 50000, 53000, 
             49000, 46000, 51000, 65000, 250000)

```

```{r}

my_data

```

Use the `mean()` function. Click the green arrow below to
"`Run Current Chunk`".

```{r}

mean(my_data)

```

*Replace this text with your interpretation of the mean*

## Checkpoint 03

Replace the `mean` function with the `median` function below.

```{r}

mean(my_data)

```

*Replace this text with your interpretation of the median.*

## Checkpoint 04

Install your first package using the function `install.packages()`.
Click the green arrow below to "`Run Current Chunk`".

```{r}
#| eval: false
#| message: false
#| label: install-carData

install.packages("carData")
```

## Checkpoint 05

Load the package into the environment using the `library()` function.
Run the code chunk below.

```{r}

library("carData")

```

See a list of the data-sets available in the carData package. Run the
code chunk below.

```{r}

data(package = "carData")

```

## Checkpoint 06

Load the `Salaries` data professors' salaries into the `R Environment`.
Run the code chunk below.

```{r}

data("Salaries")

```

Take a look at the data by clicking on the `Salaries` object in the
`R Environment`.

How many professors are in this dataset? How many variables? What are
the variable names?

## Checkpoint 07

Calculate and interpret the mean salary for professors in this dataset.

```{r}

mean(Salaries$salary)

```

Calculate the median salary for professors in this dataset. What does
the difference between the mean and median suggest about the data?

```{r}

median(Salaries$salary)

```

What's the average number of years since receiving a PhD for professors
in this sample? What's the average years of service? (*Notice there is
one line per function.*)

```{r}

mean(Salaries$yrs.since.phd)
mean(Salaries$yrs.service)

```

## Checkpoint 08

On average, how much more money do the men make compared with women?

```{r}

tapply(Salaries$salary, Salaries$sex, mean)

```

Which discipline earns the most money, on average?

```{r}

tapply(Salaries$salary, Salaries$discipline, mean)

```

## Checkpoint 09

What's the modal rank for these professors?

```{r}

table(Salaries$rank)

```

## Checkpoint 10

```{r}

summary(Salaries)

```

# Bonus Rounds

1.  Choose any of the other available data sets from the `carData`
    package and use the functions you've learned to say something
    interesting about the data. Suggested data sets: `SLID` and
    `Migration`.

    *Add multiple lines of code to a code chunk or add code chunks as
    necessary.*

```{r}



```

2.  Click the `Render` button above this quarto document and view the
    output.
