---
title: "Milestone 02"
author: "Answer Key"
format: pdf # or replace 'pdf' with 'typst'
execute:
  error: true # change this to false when your assignment is complete.
---

# Environment

Update the YAML with your first and last name.

Load the packages.

```{r}
#| label: load
#| message: false

library(tidyverse) 
library(haven) 
library(gssr)
library(gssrdoc)
library(summarytools)
```

Load the 2024 GSS data and the GSS panel data

```{r}
#| label: gss-data
#| message: false

# ADD YOUR CODE HERE TO LOAD THE GSS 2024 DATA 
gss24 <- gss_get_yr(2024)

```

You’ll be working with the following GSS variables:

-   `hrs1`: Hours worked per week
-   `tvhours`: Hours spent watching TV per day
-   `sex`: Respondent’s gender
-   `polviews`: Political views (from extremely liberal to extremely
    conservative)

# Code, Output, Meaning

*Use R to complete the checkpoints below. Show your work (e.g., R code
chunks) where appropriate. Add narrative (text outside code chunks) or
comments (text inside the code chunks) throughout.*

Reference specific statistics (where appropriate) from your output to
justify your answers. Explain what the values tell you about the data;
interpret their meaning in relation to the question.

## Checkpoint 01: Clean variables

Use `zap_missing()`, `as_factor()`, and `droplevels()` to make `sex`
into a clean factor variable.

```{r}
#| label: sex-01

gss24$sex <- zap_missing(gss24$sex)
gss24$sex <- as_factor(gss24$sex)
gss24$sex <- droplevels(gss24$sex)
```

## Checkpoint 02: Summarize weekly work hours (`hrs1`)

Use `drop_na` and `sumamrise()` to create a table showing the frequency,
median, mean, and sd of **`hrs1`**

**BONUS**: use `round()` to indicate keeping 2 decimal places.

```{r}
#| label: table-01

gss24 |>
  drop_na(hrs1) |>
  dplyr::summarize(
    freq = n(),
    median_hrs1 = round(median(hrs1), digits = 2),
    mean_hrs1 = round(mean(hrs1), digits = 2), 
    sd_hrs1 = round(sd(hrs1), digits = 2)
    ) 
```

***Is the median value for `hrs1` consistent with the mean? Speculate
what might explain any differences.***

The median (40) and mean (39.44) are close, suggesting little skew. The slightly lower mean than median suggests a mild left (e.g., negative) skew, indicating a few respondents report very low hours (likely due to unemployment or few part-time hours).

**Interpret the mean and standard deviation for `hrs1`.**

On average, people in the sample work just under a standard full-time schedule of 40 hours per week (39 hours). The standard deviation of 14 hours indicates 68% of respondents report between 25.6 and and 53 weekly work hours. This is a substantial amount of variation, as some people work significantly fewer or more hours per week. This likely reflects a mix of part-time, full-time, and overtime work patterns.

## Checkpoint 03: Summarize dailty tv hours (`tvhours`)

Use `drop_na()` and `summarize()` to create a table showing the
frequency, median, mean, and sd of **`tvhours`**.

**BONUS**: use `round()` to indicate keeping 2 decimal places.

```{r}
#| label: table-02

gss24 |>
  drop_na(tvhours) |>
  summarize(
    freq = n(),
    median_tv = round(median(tvhours), digits = 2),
    mean_tv = round(mean(tvhours), digits = 2), 
    sd_tv = round(sd(tvhours), digits = 2)
    ) 
```

***Is the median value for `tvhours` consistent with the mean? Speculate
what might explain any differences.***

**Interpret the mean and standard deviation for `tvhours`.**

## Checkpoint 04: Summarize daily tv hours (`tvhours`) by `sex`

Use `drop_na`, `group_by()` and `sumamrize()` to create a table showing
the mean and sd of `tvhours` by `sex`.

```{r}
#| label: table-03

gss24 |>  
  drop_na(tvhours, sex)  |>
  group_by(sex) |>  
  summarize(
    mean = round(mean(tvhours), digits = 2), 
    sd = round(sd(tvhours), digits = 2)
    ) 

```

**How did the mean and standard deviation of daily tv hours differ
between men and women and what do these differences suggest about
gender-based variability in tv viewing patterns?**

## Checkpoint 05: Create a new data frame

Create a new data frame called `my_data` that contains only the four
variables listed above. (Hint: use `select()`).

Then, use `head()` to preview `my_data`.

```{r}
#| label: select

my_data <- gss24 |>
  select(hrs1, tvhours, sex, polviews) 

head(my_data)
```

## Checkpoint 06: Remove rows with missing data

Use `drop_na()` to keep only respondents who have non-missing values for
all variables in `my_data`.

Then, use `head()` to preview `my_data`.

```{r}
#| label: drop-na

my_data <- my_data |>
  drop_na()

head(my_data)
```

## Checkpoint 07: Create variables

A.  Use `mutate()` and `case_when()` to add a new variable to `my_data`
    called `work_category` that groups respondents as follows:

-   "Part-time" if hrs1 is equal to or more than 0 and equal to or less
    than 34
-   "Full-time" if `hrs1` is 35 or more

**BONUS**: use `table()` to show the counts of your new variable.

```{r}
#| label: work_category

my_data <- my_data |>
  dplyr::mutate(
    work_category = case_when(
      hrs1 >= 0 & hrs1 <= 34 ~ "Part-time",
      hrs1 >= 35 ~ "Full-time"
  ))

# table(my_data$hrs1, my_data$work_category)
table(my_data$work_category)
```

B.  Use `mutate()` and `case_when()` to add a new variable to `my_data`
    called `pol3cat` that groups respondents as follows:

-   "Liberal" if `polviews` equals "extremely liberal", "liberal", or
    "slightly liberal"
-   "Moderate" if `polviews` equals "moderate, middle of the road"
-   "Conservative" if `polviews` equals "extremely conservative",
    "conservative", "slightly conservative"

**BONUS**: use `table()` to show the cross-tab of `pol3cat` and
`polviews`

```{r}
#| label: pol3cat

my_data <- my_data |>
  dplyr::mutate(
    pol3cat = case_when(
      polviews >= 1 & polviews <= 3 ~ "Liberal",
      polviews == 4 ~ "Moderate",
      polviews >= 5 & polviews <= 7 ~ "Conservative"
  )) 

table(my_data$polviews, my_data$pol3cat)
```

## Checkpoint 08: Summarize tv hours by work category

Use `group_by()` and `sumamrize()` to create a table showing the
frequency, mean, median, and sd of **`tvhours`** for each of the two
work-hours groups.

**BONUS**: use `round()` to indicate keeping 2 decimal places.

```{r}
#| label: table-04

my_data |>
  group_by(work_category) |>  
  dplyr::summarize(
    count = n(),
    mean_tvhours = round(mean(tvhours), digits = 2), 
    median_tvhours = round(median(tvhours)),
    sd_tvhours = round(sd(tvhours), digits = 2)
    ) 
```

***Which work-hours group reported the most daily tv hours in 2024? Was
there a lot or a little variability in tv hours?***

***Is the median value for `tvhours` consistent with the mean? Speculate
what might explain any differences.***

## Checkpoint 09: Summarize weekly work and daily tv hours by political identity

A.  Use `group_by()` and `sumamrize()` to create a table showing the
    frequency, mean, median, and sd of **`hrs1`** for each of the three
    political identity groups.

**BONUS**: use `round()` to indicate keeping 2 decimal places.

```{r}
#| label: table-05

my_data |>
  group_by(pol3cat) |>  
  dplyr::summarize(
    count = n(),
    mean_hrs1 = round(mean(hrs1), digits = 2), 
    median_hrs1 = round(median(hrs1)),
    sd_hrs1 = round(sd(hrs1), digits = 2)
    ) 
```

***Which political group reported the most work hours in 2024? Was there
a lot or a little variability in work hours?***

***Is the median value for*** **`hrs1`** ***consistent with the mean?
Speculate what might explain any differences.***

B.  Use `group_by()` and `sumamrize()` to create a table showing the
    frequency, mean, median, and sd of **`tvhours`** for each of the
    three political identity groups. **BONUS**: use `round()` to
    indicate keeping 2 decimal places.

```{r}
#| label: table-06

my_data |>
  group_by(pol3cat) |>  
  dplyr::summarize(
    count = n(),
    mean_tv = round(mean(tvhours), digits = 2), 
    median_tv = round(median(tvhours)),
    sd_tv = round(sd(tvhours), digits = 2)
    ) 
```

***Which political group watched the most television in 2024? Was there
a lot or a little variability in television hours?***

## Checkpoint 10: Create a summary dataframe

Use `group_by()` and `summarize()` to create and save a dataframe
showing the average work (name it `avg_work_24` and tv hours (name it
`avg_tv_24`) for men and women in 2024.

```{r}
#| label: summary

gss24_summary <- my_data |>
  group_by(sex) |>  
  dplyr::summarize(
    avg_work_24 = round(mean(hrs1), digits = 2), 
    avg_tv_24 = round(mean(tvhours), digits = 2)
    )

head(gss24_summary)
```

***Did men or women report more average work hours in 2024? What about
TV hours?***

***What percentage of men worked 40 hours or less than per week in 2024?
(You can either add `R` code chunk to this document or calculate it with
the formula. Show your work either way.)***

```{r}
#| label: percentage

my_data|>
  dplyr::mutate(extra_hrs = case_when(
    hrs1 <= 40 ~ 1,
    hrs1 >  40 ~ 0,
    TRUE ~ NA_integer_)) |>
  group_by(sex) |>
  summarise(proportion = mean(extra_hrs)
)

```

# IPUMS Data

To keep your Research Brief progress on track, you'll complete short
exercises that correspond with the new course material using your own
dataset as part of your milestones.

-   Present a (pretty) relative frequency table of a nominal or ordinal
    variable in your dataset.

-   Create a summary table for a key interval-ratio (e.g., continuous)
    variable you are interested in exploring more in your dataset.
    Include the frequency, mean, median, and sd.

***Some potentially helpful resources:***

-   [package `ipumsr()`](https://tech.popdata.org/ipumsr/)

-   [Webinar: Using IPUMS data in R with
    ipumsr](https://www.youtube.com/live/OT6upQ1dBgU)

-   [Just the slides: Using IPUMS data in R with
    ipumsr](https://assets.ipums.org/_files/webinars/slides/ipumsr10-12-21.pdf)
