---
title: "Milestone 01"
author: "FirstName LastName"
format: pdf
---

# Environment

Update the YAML with your first and last name.

Load the packages from the code-along. (Install first if not already done so.)

```{r}
#| label: load

library(here)
library(tidyverse)
library(gssr)
library(gssrdoc)
```

# Add Documentation

Run the code and add **narrative** (outside code chunks) or **comments** (inside the code chunks) to the R chunks below.

```{r}
#| label: here

here()
```

```{r}
#| label: list-files

list.files(path = here())

list.files(path = here("project"))
```

```{r}
#| label: time

Sys.time()
```

# Explore the GSS

Load all years of the gss data. Then, use the `gss_get_yr()` command to load the data only for the 2024 survey respondents.

*Hint*: Use the code from code-along-01.qmd

```{r}
#| label: gss-datasets

# Load the data (will appear in your Global Environment pane)
data(gss_all)
 
# load the gss 2024 data (add your code below)

```

Use R to answer the questions below. Show your work (e.g., R code) when appropriate. Add narration and comments throughout.

-   Report the survey question that corresponds to the `fefam` variable.

-   What type is the `fefam` variable? (nominal, ordinal, interval-ratio, dichotomous)? Justify your answer.

-   Which years of the survey did the `fefam` variable appear?

-   What value corresponds with "agree" for this variable?

-   How many people "strongly agree" with the `fefam` statement in 2024?

-   Look at the `happy` variable. Were most people "not too happy," "pretty happy," or "very happy" in 2024?

-   Create a crosstab of the `happy` and `sex` variables using only the 2024 survey data. How many women were `very happy`? How many men were `very happy`?

# Meet the IPUMS Data

To keep your Research Brief progress on track, you'll complete short exercises that correspond with the new course material using your own dataset as part of your milestones.

Consider a research topic you are interested in and explore the documentation for some of the available [IPUMS (microdata) datasets](https://www.ipums.org/overview) that could be relevant for the subject. Start broadly, such as looking at the [Topical Availability Summary](https://www.ipums.org/topics). Then, look at a couple of individual datasets more in depth. Explore the sidebar of available documentation for a couple of datasets: About, Data, Documentation (including User Guides and Questionnaires, FAQs).

-   Write 1-2 sentences about your research interests.

-   Then, write 2-3 sentences about which IPUMS dataset you are thinking of using, explaining the sample population and time frame of the dataset and why it is a good fit for your research interests.

-   Describe 2-3 key variables available in the dataset that are a good fit for your research topic. Report the exact survey questions and response options for these variables.

-   Finally, create an IPUMS account for the dataset of interest. See this [help video](https://www.youtube.com/watch?v=Oo7RYMV7q4Y) for details.

**NOTE:** For your Research Briefs, you must choose an [IPUMS (microdata) dataset](https://www.ipums.org/overview) to use.
