---
title: "Milestone 03"
author: "FirstName LastName"
format: pdf # or replace 'pdf' with 'typst'
execute:
  error: true # change this to false when your assignment is complete.
---

# Environment

Update the YAML with your first and last name.

Load the packages.

```{r}
#| label: load
#| message: false

library(tidyverse) 
library(haven) 
library(gssr)
library(gssrdoc)
library(summarytools)
library(Rmisc) # load the new package
library(gtsummary) # load the new package
```

Load the 2022 GSS data

```{r}
#| label: gss-data

# ADD YOUR CODE HERE TO LOAD THE GSS 2022 DATA 

```

You’ll be working with the following GSS variables:

-   `meovrwrk`: Men hurt family when focus on work too much
-   `sex`: Respondent’s gender
-   `lifenow` R's rating of life overall now from 0-10
-   `educ` Highest year of school completed

# Code, Output, Meaning

*Use R to complete the checkpoints below. Show your work (e.g., R code
chunks) where appropriate. Add narrative (text outside code chunks) or
comments (text inside the code chunks) throughout.*

***Show your df and variable changes as you go (e.g., `table()`,
`head()`).*** Reference specific statistics (where appropriate) from
your output to justify your answers. Explain what the values tell you
about the data; interpret their meaning in relation to the question.

## Checkpoint 01: Create a new dataframe

A.  Create a df named `my_data` with `meovrwrk`, `sex`, `marital`,
    `lifenow`, and `educ`. Then, use `drop_na()` to remove rows with
    missing data.

```{r}
#| label: my_data

# ADD YOUR CODE HERE

```

## Checkpoint 02: Clean and create variables

A.  Use `mutate()` and `case_when()` to add a new variable to `my_data`
    called `hurt_num` that groups respondents as follows:

-   1 if `meovrwrk` is 'strongly agree' or 'agree'
-   0 if `meovrwrk` is equal to anything else

```{r}
#| label: hurtovrwrk

# ADD YOUR CODE HERE

```

B.  Use `mutate()` and `case_when()` to add a new variable to `my_data`
    called `college` that groups respondents as follows:

-   "No college degree" if `educ` is less than 16
-   "College degree" if `educ` 16 or more

```{r}
#| label: college

# ADD YOUR CODE HERE

```

C.  Use `mutate()` and `case_when()` to add a new variable to `my_data`
    called `edu_category` that groups respondents as follows:

-   "High school or less" if `educ` is 12 or less
-   "Some college" if `educ` is equal to or greater than 13 and equal to
    or less than 15
-   "College or more" if `educ` is 16 or more

**BONUS: Then, use `factor()` to put the categories into a meaningful
order**

```{r}
#| label: edu_category

# ADD YOUR CODE HERE

```

D. Use `mutate()` and `case_when()` to add a new variable to `my_data`
called `marstat` that groups respondents as follows:

-   "Married" if `marital` is 'married'
-   "Single" if `marital` is 'never married'
-   "Prev married" if `marital` is 'widowed', 'divorced', or 'separated'

**BONUS: Then, use `factor()` to put the categories into a meaningful
order**

```{r}
#| label: marstat

# ADD YOUR CODE HERE

```

E. Use `zap_missing()`, `as_factor()`, and `droplevels()` to make `sex`
and `meovrwkr` into clean factor variables.

```{r}
#| label: sex-meovrwkr

# ADD YOUR CODE HERE

```

## Checkpoint 03: Create and interpret confidence interval

A.  Calculate the mean and 95% confidence interval for `lifenow`

```{r}
#| label: ci-lifenow-95

# ADD YOUR CODE HERE

```

***Interpret your confidence interval (e.g., say in plain language what
the CI is telling us).***

B.  Calculate the mean and 99% confidence interval for `lifenow`

```{r}
#| label: ci-lifenow-99

# ADD YOUR CODE HERE

```

***How does the 95% confidence interval differ from the 99% confidence
interval in terms of width and certainty?***

## Checkpoint 04: Create, compare and interpret confidence intervals

A.  Calculate the mean and 95% confidence intervals for `lifenow` by
    `college`

```{r}
#| label: ci-lifenow-college

# ADD YOUR CODE HERE

```

***Do the confidence intervals for these two groups overlap? What does
that imply about the statistical significance of their difference?***

B. Calculate the mean and 95% confidence intervals for `lifenow` by
`edu_category`

```{r}
#| label: ci-lifenow-edu_category

# ADD YOUR CODE HERE

```

***Do the confidence intervals for these three groups overlap? What does
that imply about the statistical significance of their difference?***

## Checkpoint 05: Create, compare and interpret confidence intervals

Calculate the mean and 95% confidence intervals for `hurt_num` by
`marstat` & `sex` (e.g, married women, married men, single women,
single men, etc.)

```{r}
#| label: ci-hurt_num-marstat

# ADD YOUR CODE HERE

```

***Assess whether gender differences within each marital status group
are likely to be statistically meaningful.***

## Checkpoint 06: Write a null and research hypothesis

***Write a null and research hypothesis for the association between the
variables `lifenow` and `college`***

## Checkpoint 07: Produce and interpret a two-sample t-test

Use `t.test()` to produce a two-sided t.test of the variables `lifenow`
and `college`

```{r}
#| label: ttest

# ADD YOUR CODE HERE

```

***Explain what the test reveals about the statistical differences in
average life satisfaction between the two groups. What do the results
suggest about the null and research hypotheses?***

## Checkpoint 08: Create cross-tabs with chi-square tests

Use `tbl_cross()` and `add_p()` to create a cross-tab with a chi-square
test for the variables `meovrwrk` and `college`

```{r}
#| label: meovrwrk-college

# ADD YOUR CODE HERE

```

***Is there a statistically significant difference in agreement that men
hurt the family when they focus on work too much by college degree?***

## Checkpoint 09: Calculate and interpret a chi-square test

Use `tbl_cross()` and `add_p()` to create a cross-tab with a chi-square
test for the variables `meovrwrk` and `sex`

```{r}
#| label: meovrwrk-sex

# ADD YOUR CODE HERE

```

***Is there a statistically significant gender difference in agreement
that men hurt the family when they focus on work too much?***

## Checkpoint 10: Produce and interpret a Pearson’s Correlation Coefficient (R)

Use `cor.test()` to test the correlation between `lifenow` & `educ`

```{r}
#| label: cor-educ01

# ADD YOUR CODE HERE

```

***Describe the strength and direction of any correlation between
education and life state.***

# IPUMS Data

To keep your Research Brief progress on track, you'll complete short
exercises that correspond with the new course material using your own
dataset as part of your milestones.

-   Choose an interval-ratio (e.g, continuous or proportion) variable
    that you wish to consider as a **dependent variable** (DV).

-   Create a summary table for this DV, including the total number of
    observations, mean, median, and mode.

-   Pick or create a dichotomous variable (e.g., gender, two racial
    categories; two age categories; two countries; two years; etc.) in
    your dataset and create confidence intervals for each group for
    your DV. Compare and interpret the confidence intervals.

-   Write a null and research hypothesis using these two variables.

-   Conduct a two sample t-test (using `t.test()`) for your DV and
    dichotomous variable. Interpret your results. How do they relate to
    your hypotheses?

-   Choose a second interval-ratio variable that you think may be
    associated with your DV.

-   Use `cor.test()` to formally test for an association between the two
    continuous variables. Interpret your results.
