---
title: "milestone-00"
author: "FirstName LastName"
format: html
editor: visual
---

# Getting Started in RStudio & R

## Checkpoint 01

In the `Console`, type:

```         
print("Hello, World!)
```

Now, click the green arrow below to `Run Current Chunk`.

```{r}
print("Hello, World")
```

## Checkpoint 02

Pretend the following values are annual incomes for 10 families.

| Family ID | Annual Income (CAD) |
|:----------|:--------------------|
| F01       | \$48,000            |
| F02       | \$52,000            |
| F03       | \$45,000            |
| F04       | \$50,000            |
| F05       | \$53,000            |
| F06       | \$49,000            |
| F07       | \$46,000            |
| F08       | \$51,000            |
| F09       | \$65,000            |
| F10       | \$250,000           |

[*Source: Totally fake data*]{style="color: $gray-300; font-size: 80%"}

Store the values in an object called `my_data`. Click the green arrow
below to `Run Current Chunk`.

```{r}

my_data <- c(48000, 52000, 45000, 50000, 53000, 
             49000, 46000, 51000, 65000, 250000)

```

View your new data object. Click the green arrow below to
`Run Current Chunk`.

```{r}

my_data

```

Use the `mean()` function. Click the green arrow below to
`Run Current Chunk`.

```{r}

mean(my_data)

```

*Replace this text with your interpretation of the mean.*

## Checkpoint 03

Replace the `mean` function with the `median` function below.

```{r}

mean(my_data)

```

*Replace this text with your interpretation of the median.*

## Checkpoint 04

Install your first package using the function `install.packages()`.
Click the green arrow below to `Run Current Chunk`.

```{r}
#| eval: false
#| message: false
#| label: install-carData

install.packages("carData")
```

## Checkpoint 05

Load the package into the environment using the `library()` function.
Run the code chunk below.

```{r}

library("carData")

```

See a list of the data-sets available in the carData package. Run the
code chunk below.

```{r}

data(package = "carData")

```

## Checkpoint 06

Load the `Salaries` data which contains information on professors'
salaries into the `R Environment`. Run the code chunk below.

```{r}

data("Salaries")

```

Take a look at the data by clicking on the `Salaries` object in the
`R Environment`.

How many professors are in this dataset? How many variables? What are
the variable names?

## Checkpoint 07

Calculate and interpret the mean salary for professors in this dataset.

```{r}

mean(Salaries$salary)

```

*Replace this text with your interpretation of the mean.*

Calculate the median salary for professors in this dataset.

```{r}

median(Salaries$salary)

```

*What does the difference between the mean and median suggest about the
data?*

What's the average number of years since receiving a PhD for professors
in this sample? What's the average years of service? (*Notice there is
one line per function.*)

```{r}

mean(Salaries$yrs.since.phd)
mean(Salaries$yrs.service)

```

## Checkpoint 08

On average, how much more money do the men make compared with women?

```{r}

tapply(Salaries$salary, Salaries$sex, mean)

```

Which discipline earns the most money, on average?

```{r}

tapply(Salaries$salary, Salaries$discipline, mean)

```

## Checkpoint 09

What's the modal rank for these professors?

```{r}

table(Salaries$rank)

```

## Checkpoint 10

Use the `summary()` function to see more than one descriptive statistic
at a time.

```{r}

summary(Salaries)

```

*Why do some variables have different statistics than the others?*

# Bonus Rounds

1.  Choose any of the other available data sets from the `carData`
    package and use the functions you've learned to say something
    interesting about the data. Suggested data sets: `SLID` and
    `Migration`.

    *Add multiple lines of code to a code chunk or add code chunks as
    necessary.*

```{r}



```

2.  Replace the author name at the top of the document with your name
    (keep quotation marks). Then, click the `Render` button above this
    quarto document and view the output.
