---
title: "Milestone 03"
author: "FirstName LastName"
format: pdf # or replace 'pdf' with 'typst'
execute:
  error: true # change this to false when your assignment is complete.
---

# Environment

Update the YAML with your first and last name.

Load the packages.

```{r}
#| label: load
#| message: false

library(here)
library(tidyverse) 
library(haven) 
library(gssr)
library(gssrdoc)
library(summarytools)
library(Rmisc) # load the new package
library(gtsummary) # load the new package
```

Load the 2022 GSS data

```{r}
#| label: gss-data
#| message: false

# ADD YOUR CODE HERE TO LOAD THE GSS 2024 DATA 
gss22 <- gss_get_yr(2022)

```

You’ll be working with the following GSS variables:

-   `id`: Respondent id
-   `meovrwrk`: Men hurt family when focus on work too much
-   `sex`: Respondent’s gender
-   `lifenow` R's rating of life overall now from 0-10
-   `year`: GSS year for this respondent

# Code, Output, Meaning

*Use R to complete the checkpoints below. Show your work (e.g., R code
chunks) where appropriate. Add narrative (text outside code chunks) or
comments (text inside the code chunks) throughout.*

Reference specific statistics (where appropriate) from your output to
justify your answers. Explain what the values tell you about the data;
interpret their meaning in relation to the question.

## Checkpoint 01: Select relevant variables

A.  Create a df with `id` and only the categorical variables listed above. Then, use `zap_missing()`, `as_factor()`, and `droplevels()` to make them into clean factor variables.

```{r}
#| label: my-cat

# Categorical Variables
my_cat <- gss22 |>
  select(id, meovrwrk, sex, marital) |>
  zap_missing() |>
  as_factor() |>
  droplevels()

```

B.  Create a df with `id` and only the continuous variables. Then, use `mutate()` and `as.numeric()` to make them numerical variables.

```{r}
#| label: my-con

# Continuous Variables
my_con <- gss22 |>
  select(id, lifenow, educ) |>
  mutate(
    lifenow = as.numeric(lifenow))
```

C.  Combine the two dataframes and name it `my_data`.

```{r}
#| label: my-data

# Combine the two dataframes
my_data <- left_join(my_cat, my_con, by = "id")
```

## Checkpoint 02: Remove rows with missing data

Use `drop_na()` to keep only respondents who have non-missing values for
all variables.

```{r}
#| label: drop-na

my_data <- my_data |>
  drop_na()

```

## Checkpoint 03: Create cross-tabs with chi-square tests

Use `tbl_cross()` and `add_p()` to create a cross-tab with a chi-square test


Is there a gender difference (`sex`) in agreement that men hurt the family when they focus on work too much (`meovrwk`)?

```{r}
#| message: false
#| eval: false # change this!
#| label: tbl-cross-meovrwrk

my_data |>
  tbl_cross(
    row = meovrwrk,
    col = sex,
    percent = "column") |>
  add_p(source_note = TRUE)
```