```{r data type, echo = FALSE, results = "hide"}

pacman::p_load(
  tidyverse,
  gssr, # load U.S. gss data
  gssrdoc, # load the gss codebook
  haven, # work with labeled data
  labelled, # use labels as values
  summarytools, # easy descriptive statistics
  Rmisc, # confidence intervals
  knitr, # pretty tables
  kableExtra, # centered tables
  conflicted # declare primary packages
)

## Address important conflicts
conflict_prefer("is.labelled", "haven")
conflict_prefer("filter", "dplyr")
conflict_scout() # Identify the conflicts

## Load the gss dataset
gss22 <- gss_get_yr(2022) # 2024 survey respondents

options(digits = 2)

```

Question
========
Using 2024 GSS data, a researcher estimated the proportion of Americans who believe in life after death. Based on the 95% confidence interval from R, which interpretation is correct?


```{r output, echo = FALSE, results = "asis", fig.pos='H'}
#| message: false

gss22$sex <- zap_missing(gss22$sex) 
gss22$sex <- as_factor(gss22$sex)   
gss22$sex <- droplevels(gss22$sex)  

gss22 |>
  drop_na(hrsrelax, sex) |>
  group_by(sex) |>
  dplyr::summarise(
    n = n(),
    mean = mean(hrsrelax),
    sd = sd(hrsrelax),
    lowCI = CI(hrsrelax, ci = 0.95)['lower'],
    hiCI  = CI(hrsrelax, ci = 0.95)['upper']
    ) |>
  kable(format = "latex", booktabs = TRUE, longtable = FALSE) |>
  kable_styling(latex_options = c("hold_position"), position = "center")

```

Answerlist
----------

-   The confidence intervals overlap, so we can’t know for sure if there is a statistical difference between the average time men and women spend on leisure activities per day.

-   The confidence intervals do not overlap so we can conclude that the average time men and women spend on leisure activities per day are statistically significantly different.

-   We cannot determine whether the null hypothesis of zero difference between the parameters will be rejected at the 0.05 level.

-   We are 95% certain that women have less time per day than men for leisure activities, likely because of their greater time spent on housework and childcare.

Meta-information
================
exname: hrsrelax 
extype: schoice 
exsolution: 1000 
exshuffle: 4
