```{r data type, echo = FALSE, results = "hide"}

pacman::p_load(
  tidyverse,
  gssr, # load U.S. gss data
  gssrdoc, # load the gss codebook
  haven, # work with labeled data
  labelled, # use labels as values
  summarytools, # easy descriptive statistics
  Rmisc, # confidence intervals
  knitr, # pretty tables
  kableExtra, # centered tables
  conflicted # declare primary packages
)

## Address important conflicts
conflict_prefer("is.labelled", "haven")
conflict_prefer("filter", "dplyr")
conflict_scout() # Identify the conflicts

## Load the gss dataset
data(gss_all)

options(digits = 2)

```

Question
========
Using the U.S. General Social Survey data, a researcher tests whether the proportions who favor a law requiring a person to obtain a police permit before they could buy a gun have changed over time. Assuming an $\alpha$ level of .05, which of these statements best summarizes the researcher’s conclusion?


```{r output, echo = FALSE, results = "asis"}
#| message: false

my_data <- gss_all |>
  filter(year == 1984 | year == 2024) |>
  select(year, gunlaw) |>
  drop_na(gunlaw)

# Use logic to recode (yes: 1 = 1; no: 2 = 0)
my_data$gunlaw <- ifelse(my_data$gunlaw == 1, 1, 0)
  
# Run the t-test
ttest_result <- t.test(my_data$gunlaw ~ my_data$year, alternative = "two.sided")

# Create a tibble with key results
tibble(
  `t-statistic` = round(ttest_result$statistic, 2),
  `df` = round(ttest_result$parameter, 2),
  `p-value` = round(ttest_result$p.value, 4),
  `95% CI Lower` = round(ttest_result$conf.int[1], 2),
  `95% CI Upper` = round(ttest_result$conf.int[2], 2)
) |>
  kable(format = "latex", booktabs = TRUE) |>
  kable_styling(latex_options = c("hold_position"), position = "center")

```

Answerlist
----------

-   There is not a statistically significant difference in support for the police permit law between 1984 and 2024, indicating that public opinion has remained stable.

-   There is a statistically significant difference in support for the police permit law between 1984 and 2024, suggesting that public opinion has changed over time.

-   The difference in support for the police permit law between 1984 and 2024 was due to sampling error and cannot be interpreted.

-   The is a statistically significant increase in support for the police permit law from 1984 to 2024, proving that stricter gun control laws are now favored by most Americans.

Meta-information
================
exname: ttest 
extype: schoice 
exsolution: 1000 
exshuffle: 4
