```{r data type, echo = FALSE, results = "hide"}

pacman::p_load(
  tidyverse,
  gssr, # load U.S. gss data
  gssrdoc, # load the gss codebook
  haven, # work with labeled data
  labelled, # use labels as values
  summarytools, # easy descriptive statistics
  Rmisc, # confidence intervals
  knitr, # pretty tables
  kableExtra, # centered tables
  conflicted # declare primary packages
)

## Address important conflicts
conflict_prefer("is.labelled", "haven")
conflict_prefer("filter", "dplyr")
conflict_scout() # Identify the conflicts

## Load the gss dataset
data(gss_all)

options(digits = 2)

```

Question
========
A researcher suspects Americans are watching more tv today than in the 1970s. Assuming an $\alpha$ level of .05, should the researcher reject the null hypothesis? Why or why not?

```{r output, echo = FALSE, results = "asis"}
#| message: false

my_data <- gss_all |>
  filter(year == 1975 | year == 2024) |>
  select(year, tvhours) |>
  drop_na(tvhours)

# Run the t-test
ttest_result <- t.test(my_data$tvhours ~ my_data$year, alternative = "two.sided")

cat("\\begin{samepage}\n")
# Create a tibble with key results
tibble(
  `t-statistic` = round(ttest_result$statistic, 2),
  `df` = round(ttest_result$parameter, 2),
  `p-value` = round(ttest_result$p.value, 4),
  `95% CI Lower` = round(ttest_result$conf.int[1], 2),
  `95% CI Upper` = round(ttest_result$conf.int[2], 2)
) |>
  kable(format = "latex", booktabs = TRUE) |>
  kable_styling(latex_options = c("hold_position"), position = "center")
cat("\\end{samepage}\n")
```

Answerlist
----------

-   Yes, because the p-value (0.005) is less than the $\alpha$ level of .05.

-   No, because the p-value (0.005) is greater than the $\alpha$ level of .05.

-   No, because the p-value (0.005) equals the $\alpha$ level of .05.

-   Yes, because the p-value (0.005) is greater than the $\alpha$ level of .05.


Meta-information
================
exname: ttest-tv 
extype: schoice 
exsolution: 1000 
exshuffle: 4
