```{r data type, echo = FALSE, results = "hide"}

pacman::p_load(
  tidyverse,
  gssr, # load U.S. gss data
  gssrdoc, # load the gss codebook
  haven, # work with labeled data
  labelled, # use labels as values
  summarytools, # easy descriptive statistics
  Rmisc, # confidence intervals
  knitr, # pretty tables
  kableExtra, # centered tables
  conflicted # declare primary packages
)

## Address important conflicts
conflict_prefer("is.labelled", "haven")
conflict_prefer("filter", "dplyr")
conflict_scout() # Identify the conflicts

## Load the gss dataset
data(gss_all)

options(digits = 2)

```

Question
========
In the 1970s, Americans watched an average of 3.05 TV hours per day, compared with an average of 3.3 hours per day in 2024. Assuming an $\alpha$ level of .05, what does the researcher conclude about the statistically significant difference?

```{r output, echo = FALSE, results = "asis", fig.pos='H'}
#| message: false

my_data <- gss_all |>
  filter(year == 1975 | year == 2024) |>
  select(year, tvhours) |>
  drop_na(tvhours)

#my_data |>
 # group_by(year) |>
  #summarise(
   # mean = mean(tvhours))

# Run the t-test
ttest_result <- t.test(my_data$tvhours ~ my_data$year, alternative = "two.sided")

cat("\\begin{samepage}\n")
# Create a tibble with key results
tibble(
  `t-statistic` = sprintf("%.2f", ttest_result$statistic),
  `df` = sprintf("%.2f", ttest_result$parameter),
  `p-value` = sprintf("%.3f", ttest_result$p.value),
  `95% CI Lower` = sprintf("%.2f", ttest_result$conf.int[1]),
  `95% CI Upper` = sprintf("%.2f", ttest_result$conf.int[2])) |>
  kable(format = "latex", booktabs = TRUE, longtable = FALSE) |>
  kable_styling(latex_options = c("hold_position"), position = "center")


cat("\\end{samepage}\n")
```

Answerlist
----------

-   The means indicate that Americans watch more TV in 2024 than in the 1970s. 

-   The negative t-statistic suggests a decrease, meaning Americans watch less TV in 2024 than in the 1970s.

-   Although there is a statistically significant difference, the direction of change is unclear.

-   The confidence interval is below zero, indicating Americans watch less TV in 2024 than in the 1970s.

Meta-information
================
exname: ttest-tv 
extype: schoice 
exsolution: 1000 
exshuffle: 4
