```{r data type, echo = FALSE, results = "hide"}

pacman::p_load(
  tidyverse,
  gssr, # load U.S. gss data
  gssrdoc, # load the gss codebook
  haven, # work with labeled data
  labelled, # use labels as values
  summarytools, # easy descriptive statistics
  Rmisc, # confidence intervals
  knitr, # pretty tables
  kableExtra, # centered tables
  conflicted # declare primary packages
)

## Address important conflicts
conflict_prefer("is.labelled", "haven")
conflict_prefer("filter", "dplyr")
conflict_scout() # Identify the conflicts

## Load the gss dataset
data(gss_all)

options(digits = 2)

```

Question
========
Using the U.S. General Social Survey data, a researcher suspects Americans are watching more tv today than in the 1970s. Assuming an $\alpha$ level of .05, should the researcher reject the null hypothesis? Why or why not?

```{r output, echo = FALSE, results = "asis"}
#| message: false

my_data <- gss_all |>
  filter(year == 1975 | year == 2024) |>
  select(year, tvhours) |>
  drop_na(tvhours)

t.test(my_data$tvhours ~ my_data$year, alternative = "two.sided")

```

Answerlist
----------

-   Yes, because the p-value (0.005) is less than the $\alpha$ level of .05.

-   No, because the p-value (0.005) is greater than the $\alpha$ level of .05.

-   No, because the p-value (0.005) equals the $\alpha$ level of .05.

-   Yes, because the p-value (0.005) is greater than the $\alpha$ level of .05.


Meta-information
================
exname: ttest-tv 
extype: schoice 
exsolution: 1000 
exshuffle: 4
