```{r data type, echo = FALSE, results = "hide"}

pacman::p_load(
  tidyverse,
  gssr, # load U.S. gss data
  gssrdoc, # load the gss codebook
  haven, # work with labeled data
  labelled, # use labels as values
  summarytools, # easy descriptive statistics
  Rmisc, # confidence intervals
  knitr, # pretty tables
  kableExtra, # centered tables
  conflicted # declare primary packages
)

## Address important conflicts
conflict_prefer("is.labelled", "haven")
conflict_prefer("filter", "dplyr")
conflict_scout() # Identify the conflicts

## Load the gss dataset
gss24 <- gss_get_yr(2024) # 2024 survey respondents

options(digits = 2)

```

Question
========
Using 2024 GSS data, a researcher estimated how many Americans believe in life after death. Based on the 90% confidence interval from R, which interpretation is correct?

```{r output, echo = FALSE, results = "asis", fig.pos='H'}
#| message: false

gss24$ghosts <- ifelse(gss24$postlife == 1, 1, 0)
  
gss24 |>
  drop_na(ghosts) |>
  dplyr::summarise(
    n = n(),
    mean = mean(ghosts),
    sd = sd(ghosts),
    lowCI = CI(ghosts, ci = 0.90)['lower'],
    hiCI  = CI(ghosts, ci = 0.90)['upper']
    ) |>
  kable(format = "latex", booktabs = TRUE, longtable = FALSE) |>
  kable_styling(latex_options = c("hold_position"), position = "center")

```

Answerlist
----------

-   The researcher is 90% confident that the proportion of U.S. adults who believe in some form of life after death is between .81 and .85.

-   If another survey had been conducted in 2024, there’s a 90% chance the point estimate would be between .81 and .85.

-   About 90% of people responded to the national survey.

-   Because the mean is within the confidence interval, there is a .90 probability that the proportion of U.S. adults who believe in some form of life after death is .83.

Meta-information
================
exname: ghosts 
extype: schoice 
exsolution: 1000 
exshuffle: 4
