```{r data type, echo = FALSE, results = "hide"}

pacman::p_load(
  tidyverse,
  gssr, # load U.S. gss data
  gssrdoc, # load the gss codebook
  haven, # work with labeled data
  labelled, # use labels as values
  summarytools, # easy descriptive statistics
  knitr, # pretty tables
  kableExtra, # centered tables
  conflicted # declare primary packages
)

## Address important conflicts
conflict_prefer("is.labelled", "haven")
conflict_prefer("filter", "dplyr")
conflict_scout() # Identify the conflicts

## Load the gss dataset
gss22 <- gss_get_yr(2022) # 2022 survey respondents

```

Question
========
A respondent reports 5 poor mental health days in the past 30 days. Based on summary stats for the `mntlhlth` variable, which of the following best describes their z-score? 

```{r output, echo = FALSE, results = "asis"}
#| message: false

gss22 |>  
  drop_na(mntlhlth) |> 
  dplyr::summarize(
    freq = n(), 
    median = round(median(mntlhlth), digits = 2), 
    mean = round(mean(mntlhlth), digits = 2),
    sd = round(sd(mntlhlth), digits = 2)
    ) |>
  kable(format = "latex", booktabs = TRUE) |>
  kable_styling(latex_options = c("hold_position"), position = "center")

```

Answerlist
----------

* The z-score is close to zero because 5 is close to the mean.

* The z-score is negative because 5 is below the sd.

* The z-score is large and positive because 5 is much higher than the mean.

* The z-score is undefined because the median is 0.

Meta-information
================
exname: z-score-02
extype: schoice
exsolution: 1000
exshuffle: 4
