```{r data type, echo = FALSE, results = "hide"}
#| message: false

pacman::p_load(
  tidyverse,
  gssr, # load U.S. gss data
  gssrdoc, # load the gss codebook
  haven, # work with labeled data
  labelled, # use labels as values
  summarytools, # easy descriptive statistics
  gtsummary, # tbl_cross
  gt, # make latex tables
  webshot,  # Needed for saving as image
  conflicted # declare primary packages
)

## Load the gss dataset
data(gss_all)

options(digits = 2)

```

Question
========
Has public opinion shifted on whether Black Americans face worse jobs, income, and housing than white Americans due to discrimination? Using a significance level of $\alpha = 0.01$, what conclusion should be drawn from the test results?

```{r output, echo = FALSE, results = "asis"}
#| message: false

my_data <- gss_all |>
  filter(year == 2022 | year == 2024) |>
  select(year, racdif1) |>
  dplyr::mutate(
    year = factor(year),
    opinion = ifelse(racdif1 == 1, "yes", "no")
  ) |>
  drop_na(opinion)

tbl_chi <- my_data |>
  tbl_cross(
    row = opinion,
    col = year,
    percent = "column",
    missing = "no"
  ) |>
  add_p(source_note = TRUE) 

flextable_obj <- as_flex_table(tbl_chi)
knitr::knit_print(flextable_obj)

```


Answerlist
----------

-   The evidence shows attitudes about racial inequality differ between 2022 and 2024.

-   There is no statistically significant difference in attitudes between 2022 and 2024.

-   The p-value of .000 means the null hypothesis is true.

-   The results show that most people in both years agreed racial inequality is due to discrimination.


Meta-information
================
exname: chi-square-output 
extype: schoice 
exsolution: 1000 
exshuffle: 4
