```{r data type, echo = FALSE, results = "hide"}

pacman::p_load(
  tidyverse,
  gssr, # load U.S. gss data
  gssrdoc, # load the gss codebook
  haven, # work with labeled data
  labelled, # use labels as values
  summarytools, # easy descriptive statistics
  Rmisc, # confidence intervals
  knitr, # pretty tables
  kableExtra, # centered tables
  conflicted # declare primary packages
)

## Address important conflicts
conflict_prefer("is.labelled", "haven")
conflict_prefer("filter", "dplyr")
conflict_scout() # Identify the conflicts

# Get the data from the 2022 survey
gss22 <- gss_get_yr(2022)

options(digits = 2)

```

Question
========
Based on the table, which description best describes the correlation between personal income (`realrinc`) and number of days of poor mental health (`mntlhlth`)?

```{r output, echo = FALSE, results = "asis", fig.pos='H'}
#| message: false

my_data <- gss22 |>
  select(realrinc, mntlhlth) |>
  drop_na()

cor_result <- cor.test(
  ~ realrinc + mntlhlth,
  data = my_data,
  method = "pearson",
  na.action = na.omit
)

# Create a tibble with selected results
tibble(
  `Correlation Coefficient` = sprintf("%.3f", cor_result$estimate),
  `t-statistic` = sprintf("%.2f", cor_result$statistic),
  `df` = sprintf("%.0f", cor_result$parameter),
  `p-value` = sprintf("%.3f", cor_result$p.value),
  `95% CI Lower` = sprintf("%.2f", cor_result$conf.int[1]),
  `95% CI Upper` = sprintf("%.2f", cor_result$conf.int[2])) |>
  kable(format = "latex", booktabs = TRUE, longtable = FALSE) |>
  kable_styling(latex_options = c("hold_position"), position = "center")



```

Answerlist
----------

-   Results indicate an inverse relationship (r = -0.11, p < .05), suggesting that the number of days of poor mental health decreases as personal income increases.

-  There is no evidence of an association between personal income and days of poor mental health (t = -5).

-   There is a weakly positive association between personal income and days of poor mental health (r = -0.11, p < .05).

-   Decreases in personal income was correlated with decreases in the number of days of poor mental health (t = -5, p < .05).


Meta-information
================
exname: cor.test 
extype: schoice 
exsolution: 1000 
exshuffle: 4
