```{r data type, echo = FALSE, results = "hide"}

pacman::p_load(
  tidyverse,
  gssr, # load U.S. gss data
  gssrdoc, # load the gss codebook
  haven, # work with labeled data
  labelled, # use labels as values
  summarytools, # easy descriptive statistics
  knitr, # pretty tables
  kableExtra, # centered tables
  conflicted # declare primary packages
)

## Address important conflicts
conflict_prefer("is.labelled", "haven")
conflict_prefer("filter", "dplyr")
conflict_scout() # Identify the conflicts

## Load the gss dataset
gss24 <- gss_get_yr(2024) # 2024 survey respondents

```

Question
========
What does the following R code produce?

```{r output, eval = FALSE, results = "asis"}
#| message: false

my_data |>
  tbl_cross(
    row = incgap,
    col = sex,
    percent = "column",
    missing = "no") |>
  add_p(source_note = TRUE)

```

Answerlist
----------

-   A cross-tabulation of `incgap` by `sex`, showing column percentages and a chi-square test of independence

-   A summary table comparing the mean values of `incgap` between men and women using a t-test

-   A regression output predicting `incgap` from `sex`, including R-squared and p-values

-   A frequency table of `incgap` responses with missing values imputed and grouped by `sex`

Meta-information
================
exname: tbl_cross 
extype: schoice 
exsolution: 1000 
exshuffle: 4
