---
title: "Lab 02"
format:
  revealjs:
    multiplex: false
    controls: true
    theme: [default, custom.scss]
    highlight-style: atom-one
    include-in-header:
    - file: mathjax-color.html
    slide-number: c
    math: mathjax
revealjs-plugins:
  - quiz
embed-resources: true
filters:
    - openlinksinnewpage
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: true
---

```{r echo=FALSE}

#| label: setup
#| message: false
#| include: false

## Load packages, custom functions, and styles
source("custom-setup-styles.R")

# Color in math equations
# https://egallic.fr/Recherche/aide-memoire/quarto_equation_colors.html#colors-in-equations-with-revealjs
```

# Agenda {data-menu-title="Agenda" background-image="images/calendar.png" background-size="100px" background-repeat="repeat"}

-   Frequency Distributions
-   Central Tendency
-   Variability

# Learning objectives

By the end of the lab, you will be able to ...

-   A
-   B
-   C

# Code-along 02

Download and open [code-along-02.qmd]()

# Frequency Distributions {.theme-section}

##  {data-menu-title="Think Like a Statistician" background-image="images/lightbulb-solid.png" background-size="80px" background-repeat="repeat"}

Political polarization is high in the U.S. these days.

Using the most recent survey data, do more liberals than conservatives
think sex before marriage is 'not wrong at all.'

<br>

[{{< fa regular lightbulb >}} How do we answer that
question?]{style="color: #F39C12; font-size: 175%; font-family: 'Shadows Into Light'"}

## Packages

Load the standard packages.

```{r}
#| message: false
#| label: load

library(here)
library(tidyverse)
library(gssr)
library(gssrdoc)
```

<br>

Install and load the
[`summarytools`](https://cran.r-project.org/web/packages/summarytools/vignettes/introduction.html)
package.

```{r}
#| message: false
#| label: install
#| eval: false

install.packages("summarytools")
```

<br>

```{r}
#| message: false
#| label: summarytools

library(summarytools)
```

## Load your data & codebook

```{r}
#| message: false
#| label: gssdata

# Get the data only for the 2024 survey respondents
gss24 <- gss_get_yr(2024)

# Load the codebook
data(gss_dict)

```

## Variable Descriptions

Let's familiarize ourselves with the `premarsx` and `polviews`
variables.  

<br>
  
In the console, type `?premarsx` and hit enter. The Help pane will show
you the question text and response options and values.  

<br>

Now, do the same for `polviews`.

## Table

Add a line below to also see a table for the `polviews` variable.

```{r}
#| label: table-premarsx

table(gss24$premarsx)
```

. . .

<br>

```{r}
#| label: table-polviews

table(gss24$polviews)
```

## Labels 

Use
[`haven::as_factor`](https://haven.tidyverse.org/reference/as_factor.html)
to see the value labels instead of the value numbers. Then, do the same for `polviews`.


```{r}
#| label: label.01-premarsx
#| output-location: column-fragment

table(as_factor(gss24$premarsx))
```


## Labels {.scrollable}

```{r}
#| label: label.01-polviews
#| output-location: column

table(as_factor(gss24$polviews))
```


## Better Labels

Let's use `zap_missing` and `as_factor` to clean these up. Then, do the
same for the `polviews` variable.

```{r}
#| label: label.02-premarsx
#| code-line-numbers: "1|2|3"
#| output-location: fragment

gss24$premarsx <- zap_missing(gss24$premarsx)
gss24$premarsx <- as_factor(gss24$premarsx)
table(gss24$premarsx)
```


## Better Labels {.scrollable}

```{r}
#| label: label.02-polviews

# polviews
gss24$polviews <- zap_missing(gss24$polviews)
gss24$polviews <- as_factor(gss24$polviews)
table(gss24$polviews)
```

## Better Labels cont.

Let's get ride of the empty levels in `premarsx`.

Then, do the same for `polviews`.

```{r}
#| label: label.03-premarsx
#| code-line-numbers: "1|2"
#| output-location: fragment

gss24$premarsx <- droplevels(gss24$premarsx)
table(gss24$premarsx)
```

## Better Labels cont.

```{r}
#| label: label.03--polviews
#| eval: false

gss24$polviews <- droplevels(gss24$polviews)
table(gss24$polviews)
```


## Frequency Table {.smaller}

Make a frequency table. One of `summarytools` main purposes is to help
cleaning and preparing data for further analysis. Pay attention to the
missing values.

```{r}
freq(gss24$premarsx) 
```

## Pretty Frequency Table

Using `report.nas = FALSE` suppresses the missing data.\
The `headings = FALSE` parameter suppresses the heading section.

<br>

```{r}
freq(gss24$premarsx, report.nas = FALSE, headings = FALSE) 
```

##  {.quiz-question data-menu-title="Knowledge Check" background-image="images/clipboard-question.png" background-size="100px" background-repeat="repeat"}

::: {style="color: #F39C12"}
{{< fa clipboard-question >}} Based on your table, what percentage of
respondents believe sex before marriage is 'almost always wrong'?
:::

-   [8.52]{.correct data-explanation="💯 Correct!"}
-   [24.71]{data-explanation="😲 This is the % who think premartial sex is 'always wrong'."}
-   [33.23]{data-explanation="😲 This is the % who think premarital sex is 'always' OR 'almost always wrong'."}
-   [19.05]{data-explanation="😲 This is % who think premarital sex is 'wrong only sometimes'."}

##  {.quiz-question data-menu-title="Knowledge Check" background-image="images/clipboard-question.png" background-size="100px" background-repeat="repeat"}

::: {style="color: #F39C12"}
{{< fa clipboard-question >}} Based on your table, what percentage of
respondents believe sex before marriage is 'always' or 'almost always
wrong'?
:::

-   [33.23]{.correct data-explanation="💯 Correct!"}
-   [24.71]{data-explanation="😲 This is the % who think premartial sex is 'always wrong'."}
-   [8.52]{data-explanation="😲 This is the % who think premarital sex is 'almost always wrong'."}
-   [19.05]{data-explanation="😲 This is % who think premarital sex is 'wrong only sometimes'."}
