```{r data type, echo = FALSE, results = "hide"}

pacman::p_load(
  tidyverse,
  gssr, # load U.S. gss data
  gssrdoc, # load the gss codebook
  haven, # work with labeled data
  labelled, # use labels as values
  summarytools, # easy descriptive statistics
  conflicted # declare primary packages
)

## Address important conflicts
conflict_prefer("is.labelled", "haven")
conflict_prefer("filter", "dplyr")
conflict_scout() # Identify the conflicts

## Load the gss dataset
gss24 <- gss_get_yr(2024) # 2024 survey respondents

# Load the codebook
data(gss_dict)

gss24$happy <- zap_missing(gss24$happy)
gss24$happy <- as_factor(gss24$happy)
gss24$happy <- droplevels(gss24$happy)

gss24$sex <- zap_missing(gss24$sex)
gss24$sex <- as_factor(gss24$sex)
gss24$sex <- droplevels(gss24$sex)

```

Question
========
Based on the table, which statement best compares the relative frequency of being 'not too happy' between men and women?

```{r output, echo = FALSE, results = "asis"}
#| message: false

ctable(gss24$happy, gss24$sex, 
  prop = "c", 
  format = "p", 
  useNA = "no",
  headings = FALSE
) 

```

Answerlist
----------

* Men have a slightly higher relative frequency of being 'not too happy' than women, even though fewer men reported it.

* A greater number of women reported being 'not too happy' than men, so women have a higher relative frequency.

* The frequency of being 'not too happy' is equal for men and women.

* Women have a slightly higher relative frequency because more women than men reported being 'not too happy'.

Meta-information
================
exname: relative-freq
extype: schoice
exsolution: 1000
exshuffle: 4
